---
title: "Inference"
subtitle: "POLSCI 630: Probability and Basic Regression"
format: clean-revealjs
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
date: 2/4/2025
embed-resources: true
---

```{r echo=FALSE, warning=FALSE}
library(car)
library(carData)
data(trees)
```

## Statistical inference

To this point, we have made claims about the expected value and variance of the OLS estimator, $\boldsymbol{\hat{\beta}}$

<br/>

To make probabilistic claims about $\boldsymbol{\beta}$ based on our sample, we need to more fully characterize the distribution of $\boldsymbol{\hat{\beta}}$

## Sampling distributions

Just like the mean estimator for a single variable has a [sampling distribution]{.alert}, so will our OLS estimator, $\hat{\beta}$.

```{r fig.width=6, fig.asp=0.5, fig.align='center'}

# take 10,000 random samples of 20 from uniform population on 0-1
hist(sapply(1:10000, function(x) mean(runif(20))), main="sampling distribution for mean")

```

## Normality assumption

Our model to this point is as follows:

$$y_i = \boldsymbol{x}_i'\boldsymbol{\beta} + u_i$$

And we've made two assumptions about $u_i$:

4.  Exogeneity: $\text{E}(u_i|\boldsymbol{x}_i) = 0$
5.  Homoskedasticity: $\space \text{Var}(u_i|\boldsymbol{x}_i) = \sigma^2$

. . .

Let's make another, more stringent assumption (call it 6.):

$$u_i \sim Normal(0,\sigma^2)$$

## Normality assumption

$$u_i \sim Normal(0,\sigma^2)$$

As previously, this implies $\text{E}(u_i|\boldsymbol{x}_i) = 0$ and $\text{Var}(u_i|\boldsymbol{x}_i) = \sigma^2$

. . .

But $u_i \sim Normal(0,\sigma^2)$ is an *additional claim*:

-   The probability density function for $u_i$ is Normal
-   This is a restrictive assumption! So what do we get in return?

## If we assume normal errors...

$\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{y}$

. . .

$= (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{u}]$

. . .

$= (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X}\boldsymbol{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{u}$

. . .

$= \mathbf{I}\boldsymbol{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{u}$

. . .

$\hat{\boldsymbol{\beta}} - \boldsymbol{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\boldsymbol{u}$

. . .

If the right-hand side is a linear function, [$(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$]{.alert}, of a normally-distributed random variable, [$\boldsymbol{u}$]{.alert}, the left-hand side (which subtracts a constant, [$\boldsymbol{\beta}$]{.alert}) is also normally distributed.

## Sampling distribution for $\boldsymbol{\hat{\beta}}$

Given these 6 assumptions, we can say that:

$$\hat{\beta}_k \sim Normal(\beta_k, \text{Var}(\hat{\beta}_k))$$

. . .

In matrix form:

$$\boldsymbol{\hat{\beta}} \sim MVNormal(\boldsymbol{\beta}, \sigma^2(\mathbf{X}'\mathbf{X})^{-1})$$

. . .

And any linear combination of the $\beta_k$ is also normally distributed

## Importance of normality assumption

In order for OLS to be [BLUE]{.alert}, we only need the assumptions we talked about last week

-   Linear in the parameters
-   Observations i.i.d.
-   No perfect collinearity
-   Exogeneity
-   Homoskedastic errors

. . .

The additional normality assumption is necessary for *statistical inference*, i.e. representing our beliefs about $\boldsymbol{\beta}$

## $\sigma^2$ vs $\hat{\sigma}^2$

We rarely know the true value of $\sigma^2$

Instead, we substitute $\hat{\sigma}^2 = \sum_N \frac{\hat{u}_i^2}{N - K - 1}$

$$\hat{\text{Var}}(\boldsymbol{\hat{\beta}}) = \hat{\sigma}^2(\mathbf{X}'\mathbf{X})^{-1}$$

</br>

. . .

The estimated variance of our estimator is a function of our independent variables and the estimated error variance

-   What does it look like?

## Standard errors

[Standard errors]{.alert} for the $\hat{\beta}_k$ are the square roots of the main diagonal of the covariance matrix, $\hat{\sigma}^2(\mathbf{X}'\mathbf{X})^{-1}$

-   This is the standard deviation of the sampling distributions for $\hat{\beta_k}$

::::: columns
::: column
```{r}
summary(lm(Volume ~ Height, data = trees))
```
:::

::: column
```{r, fig.height=8}
curve(dt(x, 29, 1.54), -5, 5, xlim=c(-2, 5))
```
:::
:::::

## Standard errors

::::: columns
::: {.column width="60%"}
```{r, echo = TRUE}
# define vars
X <- as.matrix(cbind(rep(1, nrow(trees)), 
                     trees[, c("Height", "Girth")]))
y <- trees$Volume

# estimate parameters
beta <- solve(t(X) %*% X) %*% t(X) %*% y

# calculate residuals
uhat <- as.numeric(y - X %*% beta)

# estimate residual variance (MSE / degrees of freedom)
s2 <- sum(uhat^2) / (nrow(X) - length(beta))

# calculate variance-covariance matrix
var_cov <- s2 * (solve(t(X) %*% X))

# standard errors
se <- sqrt(diag(var_cov))

# print coefs and SEs
round(cbind(beta, se), 4)
```
:::

::: {.column width="40%"}
```{r, echo = TRUE}
# using lm
m_trees <- lm(Volume ~ Height + Girth, data = trees)
m_trees_sum <- summary(m_trees)

# print coefs and SEs from summary object
round(m_trees_sum$coefficients[,c(1,2)], 4)
```
:::
:::::

## Sampling distribution using $\hat{\sigma}^2$

If we substitute $\hat{\sigma}^2$ for $\sigma^2$, the sampling distribution for $\hat{\boldsymbol{\beta}}$ is no longer normal

-   However, the following is true:

$$\frac{\hat{\beta}_k - \beta_k}{\hat{se}(\hat{\beta}_k)} \sim t_{(N-K-1)}$$

## t(2) vs normal distribution

```{r echo=FALSE}
par(mar=c(5,4,1,1))
curve(dt(x, 2), -5, 5, ylim=c(0,0.4))
curve(dnorm(x, 0, 1), -5, 5, add=T, lty=2)
legend("topright", c("t(2)","normal(0,1)"), lty=1:2, bty="n")
```

## Null hypothesis tests

Given that we know the CDF of the $t_{N-K-1}$ distribution, we can evaluate how extreme our estimated $\hat{\beta}_k$ is under the [null hypothesis]{.alert} that $\beta_k = 0$.

```{r, fig.height=6, fig.align='center'}
curve(dt(x, m_trees_sum$df[2]), -4, 4, xlab=expression(paste("Values of ", beta[1], " / se(", beta[1], ")")))
abline(v=c(qt(0.025, m_trees_sum$df[2]), qt(0.975, m_trees_sum$df[2])), lty=3)
text(-2.5, 0.3, "2.5 percentile")
text(2.5, 0.3, "97.5 percentile")
```

## Null hypothesis tests

Once we know $\hat{\beta}_k$ and its standard error, we can divide the former by the latter to get a:

::: incremental
1.  t-statistic: "How many standard errors away from zero is $\hat{\beta}_k$?"

2.  p-value: "How frequently would our estimate be this many standard errors from zero, under repeated sampling, if $\beta_k = 0$?"
:::

. . .

By default, political scientists tend to use two-tailed tests

-   "How frequently would our estimate be this extreme, in *either direction*"?
-   One-tailed: ...this extreme in a *particular* direction

## Null hypothesis tests

-   let you test the null hypotheses (ie, the coefficient on the predictor0 is statistically indistinguishable from 0.

-   lets you check how far away the estimated $\hat\beta_1$ is from 0 in *either* direction – and how likely that is under the assumption the true value is 0. Reject if too unlikely (p-value \<0.05 is typical).

```{r t_byhand, echo = TRUE}
# continuing the above code

# the t-statistic standardizes the coefficient
tstats <- beta / se

# the p value characterizes how extreme the t statistic is under the null
pvals <- ( 1 - pt(abs(tstats), df = nrow(X) - ncol(X)) ) * 2

reg_table <- data.frame(estimate = beta, 
                        stderr = se,
                        tstat = tstats, 
                        pval = pvals)

round(reg_table, 4)
```

## Null hypothesis tests

```{r, fig.height=8, fig.align='center'}
curve(dt(x, m_trees_sum$df[2]), -4, 4, xlab=expression(paste("Values of ", beta[1])))
abline(v=c(qt(0.025, m_trees_sum$df[2]), qt(0.975, m_trees_sum$df[2])), lty=3)
text(-2, 0.3, "2.5 percentile")
text(2, 0.3, "97.5 percentile")
abline(v=m_trees_sum$coefficients[2] / sqrt(diag(vcov(m_trees)))[2], col="red")
text(2.5, 0.2, expression(paste(beta[1])), col="red")
abline(v=-m_trees_sum$coefficients[2] / sqrt(diag(vcov(m_trees)))[2], col="red")
text(-2.5, 0.2, expression(paste("-", beta[1])), col="red")
```

## Example

```{r}
library(carData)
summary(BEPS)
```

## Example

```{r}
m_beps <- lm(Europe ~ age + gender + economic.cond.national + economic.cond.household + political.knowledge, data = BEPS)
summary(m_beps)
```

## Testing against another value {.scrollable}

Sometimes we may wish to test a null hypothesis that $\beta_k = b$, where $b$ is some specific value not equal to 0 - this is easy!

-   just subtract out the value of interest ($b$), and you are back to a standard null hypothesis test

```{r, echo = TRUE}
# test whether polknow coef is diff from -0.5
B <- coef(m_beps)["political.knowledge"]
se <- sqrt(diag(vcov(m_beps)))["political.knowledge"]
df <- m_beps$df.residual
test_value <- -0.5

# the t-statistic standardizes the coefficient
tstat <- (B - test_value) / se

# the p value characterizes how extreme the t statistic is under the null
pval <- (1-pt(abs(tstat), df = df))*2

table <- data.frame(estimate = B, 
                    stderr = se,
                    tstat = tstat, 
                    pval = pval)

round(table, 4)
```

## Confidence intervals

Thinking back to last semester: what's the frequentist interpretation of a confidence interval?

. . .

Confidence interval is a statement about the *estimator*

-   Under repeated sampling, this interval will contain the population parameter X% of the time

(When assumptions underpinning the estimator are met!)

## Lower and upper bounds

For an X% confidence interval:

::: incremental
-   The upper bound is the value of $\beta_k$ such that estimates greater than $\hat{\beta}_k$ would occur $\frac{100-X}{2}$% of the time
-   The lower bound is the value of $\beta_k$ such that estimates less than $\hat{\beta}_k$ would occur $\frac{100 - X}{2}$% of the time
-   These are just the respective quantiles of the relevant t distribution centered on $\hat{\beta}_k$, multiplied by the SE
:::

## Calculating 95% CIs

```{r echo=TRUE}
# from our previous example
B <- coef(m_beps)
df <- m_beps$df.residual
se <- sqrt(diag(vcov(m_beps)))

# calculate 95% CI for age coef
B["age"] + qt(c(0.025, 0.975), df) * se["age"]

# compare
confint(m_beps)
```

## Talking about CIs

A lot is made of the distinction between Bayesian and frequentist confidence intervals

-   freq: % of time interval will capture true value
-   Bayesian: confidence that interval contains true parameter

. . .

[(OPINION)]{.alert} I find this to be a bit silly...

-   The point of inference is to make claims about beliefs
-   The argument against probabilistic statements for frequentist seems weak
-   Frequentist interval is essentially Bayesian interval with uninformative priors

## Example

```{r echo=TRUE}
# brms = Bayesian regression modeling with stan
#library(brms)

# estimate trees with lm
confint(lm(Volume ~ Height + Girth, trees))

# estimate trees with brms (uninformative priors)
#summary(brm(Volume ~ Height + Girth, trees))

# Regression Coefficients from brms:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept   -57.91      8.97   -75.57   -40.63 1.00     3867     3303
# Height        0.34      0.14     0.07     0.60 1.00     3284     2584
# Girth         4.71      0.28     4.15     5.25 1.00     3127     2482

```

## Testing for differences between parameters

The above tests null hypotheses with respect to specific coefficients, which is *often* but not *always* what we're interested in doing.

. . .

Example: $y_i = \beta_0 + \beta_1x_{1i} + \beta_2x_{2i} + \beta_3x_{3i} + u_i$

-   $H_0: \beta_1 = \beta_2$
-   $H_1: \beta_1 > \beta_2$

## Testing for differences between parameters

Why *can't* we do the following, using $\hat{\beta_2}$ as the test value?

$$
\frac{\hat{\beta}_1 - \hat{\beta}_2}{se(\hat{\beta}_1)}
$$

. . .

There is uncertainty in *both* $\hat{\beta}_1$ *and* $\hat{\beta}_2$

-   using $\hat{\beta}_2$ as the test value is like assuming we know its true value for sure

-   instead, we need to calculate a SE based on the *combined* variance of both estimators

## Testing for differences between parameters

Example: $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + u$

-   $H_0: \beta_1 = \beta_2 \equiv \beta_1 - \beta_2 = 0$
-   $H_1: \beta_1 > \beta_2 \equiv \beta_1 - \beta_2 > 0$

. . .

Our test statistic is then:

$$
t = \frac{\hat{\beta}_1 - \hat{\beta}_2}{\sqrt{\text{Var}(\hat{\beta}_1 - \hat{\beta}_2)}}
$$

</br>

-   where, $\text{Var}(\hat{\beta}_1 - \hat{\beta}_2) = \text{Var}(\hat{\beta}_1) + \text{Var}(\hat{\beta}_2) - 2\text{Cov}(\hat{\beta}_1, \hat{\beta}_2)$

## Example

Is the coefficient for national economic conditions significantly different from the coefficient for household economic conditions?

```{r}
summary(m_beps)
```

## Example

</br>

```{r echo=TRUE}
# get relevant variances and covariances from var-cov matrix of coefficients
var_N <- vcov(m_beps)["economic.cond.national","economic.cond.national"]
var_H <- vcov(m_beps)["economic.cond.household","economic.cond.household"]
cov_NH <- vcov(m_beps)["economic.cond.national","economic.cond.household"]

# var and standard error of linear combination
var_NH <- var_N + var_H - 2*cov_NH
se_NH <- sqrt(var_NH)

# calculate test stat
( coef(m_beps)["economic.cond.national"] - coef(m_beps)["economic.cond.household"] ) / se_NH
```

## Short-cut for this simple case

```{r echo=TRUE}
m_beps_NH <- lm(Europe ~ age + gender 
                + economic.cond.national 
                + I(economic.cond.national + economic.cond.household) 
                + political.knowledge, data = BEPS)
summary(m_beps_NH)
```

## Joint hypotheses

A joint null claims that some *set* of parameters are *all* zero (or some test value)

-   for example:

$$
\beta_1 = \beta_2, ..., = \beta_K = 0
$$

-   Rejecting the null tells you *only* about the significance of the *set*, not the individual parameters

## Joint hypotheses

Why can't we just take sets of parameters out and check whether/how fit changes?

. . .

-   Removing parameters *always* decreases model fit in-sample

-   Question is whether model fit decreases *more* than would be expected simply by adding degrees of freedom

## F statistic

An [F stat]{.alert} compares residual SS of [restricted]{.alert} and [unrestricted]{.alert} models

$$F = \frac{(SSR_{r} - SSR_{ur}) / q}{SSR_{ur} / (N - K - 1)}$$

::: incremental
-   $F \sim F(q, N-K-1)$, where $q$ is \# restrictions

    -   the F distribution is formed from ratios of $\chi^2$ distributions divided by degrees freedom
    -   Because $\chi^2$ distributions are always positive, F distributions will be too
:::

## Example

Compare intercept-only model to model with all predictors (both models must be estimated w/ same data! Beware list-wise deletion!)

```{r echo=TRUE}
# estimate restricted and unrestricted models
m1_r <- lm(Europe ~ 1, data = m_beps$model)
m1_ur <- m_beps

# summary
summary(m1_ur)
```

## F test by hand

```{r echo=TRUE}
# difference in sum of squared residuals
SS_diff <- sum(m1_r$residuals^2) - sum(m1_ur$residuals^2)

# difference in degrees of freedom
df_diff <- m1_r$df.residual - m1_ur$df.residual

# SS resid for unrestricted model
SS_ur <- sum(m1_ur$residuals^2)

# degrees of freedom (unrestricted)
df_ur <- m1_ur$df.residual 

# calculate F stat
Fstat <- (SS_diff/df_diff) / (SS_ur/df_ur)

# p-value
cbind(F = Fstat, p = 1 - pf(Fstat, df_diff, df_ur))
```

## F test using `anova`

`stats::anova()` will do this for you

```{r echo=TRUE}
anova(m1_r, m1_ur)
```

. . .

This lets us say that all predictors are *jointly significant* -- still need to interpret unrestricted model to make claims about any predictor on its own

## `lm` default F test

```{r}
summary(m_beps)
```

## For subsets of predictors

```{r echo=TRUE}
anova( update(m_beps, . ~ . - age - gender, data = m_beps$model),
       m_beps
)
```

-   F test is significant, but coefficient for `gender` is not!
-   Again, this is a *joint test*!

## `car` package's `linearHypothesis`

The `car` package has a convenient function for testing linear combinations of OLS coefficients

-   syntax puts an `lm` model object in first slot, and a linear combination in the second slot (assumes 0 if no explicit equality provided)

```{r echo=TRUE}
linearHypothesis(m_beps, "age = 0")
```

## `car` package's `linearHypothesis`

```{r echo=TRUE}
linearHypothesis(m_beps, c("age = 0", "gendermale=0"))
```

## `car` package's `linearHypothesis`

```{r echo=TRUE}
linearHypothesis(m_beps, "economic.cond.national = -0.5")
```

## `car` package's `linearHypothesis`

```{r echo=TRUE}
linearHypothesis(m_beps, "economic.cond.national - economic.cond.household = 0")
```

## `car` package's `linearHypothesis`

```{r echo=TRUE}
linearHypothesis(m_beps, "economic.cond.national - 2*economic.cond.household = 0")
```

## Reminder! Need same observations

These tests *require* using the *same* observations for both restricted and unrestricted models

::: incremental
-   Invalid otherwise
-   Matters for data with missing values (listwise deletion is the default with most software)
-   When missing data is possible, and you *add* variables (as you do in these kinds of tests), there is a decent chance you will lose data
:::

## Thoughts on null hypothesis testing [(OPINION)]{.alert}

Null hypothesis testing is the norm, but it shouldn't be

-   Do we ever actually believe $\beta_k = 0$?

-   If not, what are we actually testing? (sample size?)

-   NHT also shifts focus away from the estimates themselves and the information we have about them

    -   e.g., should our interpretation drastically change when $p$ changes from 0.051 to 0.049?

## A better approach [(OPINION)]{.alert}

It is better, IMHO, to focus on:

-   Interpretation of the substantive meaning of the coefficient estimates
-   Our uncertainty in those estimates, typically characterized by a confidence interval (or two)
-   Even insignificant coefficients provide information! Just be precise about how much
