---
title: "Lab 1"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

# Formatting

-   Use hash tags to create headings (\# Problem 1) and subheadings (\## Part a)
-   `echo = F`: the output (e.g. a plot) will appear in the knitted pdf, but the code will not
-   `echo = T`: the output (e.g. a plot) and the code will appear in the knitted pdf
-   `include = F`: neither the code nor the output will appear in the knitted pdf, but the code will be evaluated
-   `fig.width = 5, fig.height = 3`: plots will appear in the knitted pdf as 5 inches wide and 3 inches tall

# Data

In this lab, we will work with data on CEO salaries from 1990, from the dataset `ceosalary.csv`. (You may remember this dataset from yesterday's lecture (slides 40 - 43), or last semester (PS 12R and Exam 3R))

The relevant variables are:

-   `salary`: 1990 compensation (\$1000s)

-   `age`: Age (years)

-   `profits`: 1990 profits (\$ millions)

```{r, include = F}
# load data
rawceosalary <- read.csv("data/ceosalary.csv")

# create new variable
rawceosalary$rawsalary <- rawceosalary$salary * 1000

# subset/reassign (ifelse and logicals) 
rawceosalary$profits <- ifelse(rawceosalary$profits < 0, NA, rawceosalary$profits) 

# remove missing (listwise deletion)
ceosalary <- na.omit(rawceosalary[, c("salary","age","profits")])
```

# Bivariate Regression

$$salary_i = \beta_0 + \beta_1 profits_{1i} + u_i$$ Find OLS estimates of $\beta_0$ and $\beta_1$ in the above model.

```{r, echo = T}
# variables 
y <- ceosalary$salary
x <- ceosalary$profits
X <- cbind(1, ceosalary$profits)

# bivariate formula
b1 <- sum((y - mean(y)) * (x - mean(x)))/sum(((x - mean(x))^2))
b0 <- mean(y) - b1 * mean(x)

b1 <- cov(ceosalary$salary, ceosalary$profits) / var(ceosalary$profits)
b0 <- mean(ceosalary$salary) - mean(ceosalary$profits)*b1

# matrix 
beta <- solve(t(X) %*% X) %*% t(X) %*% y 

# lm function
coef(lm(ceosalary$salary ~ ceosalary$profits))
```

How much is a CEO who made 0 in firm profits predicted to make in salary? How much is a CEO who made 10-million in firm profits predicted to make in salary? If firm profits increase by 1-million, what is the predicted increase in salary?

# Scatter Plot

Create a scatter plot depicting the relationship between `salary` and `profits` (with `profits` on the x-axis).

```{r, echo=F}
par(mar=c(5,4,1,1))

# scatter plot 
plot(ceosalary$profits, ceosalary$salary, pch=19,
     xlab="Firm Profits (million USD)", ylab="CEO Salary (thousand USD)",
     main = "CEO Salary by Firm Profits")

?abline() # regression line, vertical/horizontal lines, col, lty
?points() # x/y coordinates, pch, col
```

# Multiple Regression

$$salary_i = \beta_0 + \beta_1 profits_{1i} + \beta_2 age_{2i} + u_i$$ Find OLS estimates of $\beta_0$, $\beta_1$, and $\beta_3$ in the above model.

```{r, echo=T}
# matrix 
X <- cbind(1, ceosalary$profits, ceosalary$age)
y <- ceosalary$salary
beta <- solve(t(X) %*% X) %*% t(X) %*% y

# function 
m1 <- lm(salary ~ profits + age, data=ceosalary)
```

# Coefficient Table

Report the coefficient estimates, standard errors, t-statistics, and two-tailed p-values, in a professional looking table.

```{r, echo=F, results='asis', warning=F}
# install.packages("knitr")
library(knitr)
library(kableExtra)

# create data frame 
coefficient_data <- data.frame(
  Coefficient = coef(m1),
  Standard_Error = summary(m1)$coefficients[, "Std. Error"],
  T_Statistic = summary(m1)$coefficients[, "t value"],
  P_Value = summary(m1)$coefficients[, "Pr(>|t|)"]
)

coefficient_table <- knitr::kable(
  coefficient_data, 
  format="html", 
  longtable=FALSE, 
  caption = "Regression Coefficient Estimates")

styled_table <- kable_styling(
  coefficient_table,
  full_width = FALSE,
  bootstrap_options = c("striped", "hover")
)

print(styled_table)
```

# Plot Predictions

Create a figure that plots the predicted value of `salary` as a function of `profits` at three different values for `age`: the 5th percentile, the 50th percentile, and 95th percentile. Use the `predict` function to do it.

```{r, echo=F}
# 5th, 50th, and 95th percentile of age
age_percentiles <- quantile(ceosalary$age, c(0.05, 0.5, 0.95))

# range of profit values 
profits_range <- seq(min(ceosalary$profits), max(ceosalary$profits), length.out = 100)

#create dataset with every combination of profit and age 
new_data <- expand.grid(profits = profits_range, age = age_percentiles)

#predict values of salary based on profit and age in new dataset 
predicted_values <- predict(m1, newdata = new_data)

#combine new dataset with predicted salary 
predicted_df <- cbind(new_data, salary = predicted_values)
```

## Base R

```{r, echo=F}
par(mar=c(5,4,1,1))

plot(NULL, xlim = range(profits_range), ylim = range(predicted_df$salary),
     xlab = "Profits", ylab = "Predicted Salary", 
     main = "Predicted Salary vs. Profits at Different Ages")

colors = c("blue", "red", "green")

# lines
for (i in seq_along(age_percentiles)) {
  subset_data <- predicted_df[predicted_df$age == age_percentiles[i], ]
  lines(subset_data$profits, subset_data$salary, col = colors[i], lwd = 2)
}

# legend
legend("topleft", legend = paste("Age =", age_percentiles), 
       col = colors, lty = 1, lwd = 2, 
       title = "Age Percentile")

```

## ggplot

```{r, echo=F}
# install.packages("ggplot2")
library(ggplot2)

# ggplot 
ggplot(predicted_df, aes(x = profits, y = salary, color = factor(age))) +
  geom_line() +
  scale_color_manual(values = c("blue", "red", "green"),
                     labels = c(paste("Age =", age_percentiles[1]), 
                                paste("Age =", age_percentiles[2]), 
                                paste("Age =", age_percentiles[3])),
                     name = "Age Percentile") +
  labs(x = "Profits", y = "Predicted Salary",
       title = "Predicted Salary vs. Profits at Different Ages") +
  theme_minimal()

```

# Model Fit

Calculate R squared, mean squared error, and adjusted R squared by hand.

$$R^2 = 1 - \frac{SSR}{SST} = 1 - \frac{\Sigma_N (y_i - \hat{y_i})^2}{\Sigma_N (y_i - \bar{y})^2}$$ $$MSE = \frac{\Sigma_N (y_i - \hat{y_i})^2}{N - K - 1} $$ $$R_{adj}^2 = 1 - \frac{MSE}{Var(y)} = 1 - \frac{\frac{\Sigma_N (y_i - \hat{y_i})^2}{N - K - 1}}{\frac{\Sigma_N (y_i - \bar{y})^2}{N - 1}}$$

```{r, echo=F}
X <- cbind(1, ceosalary$profits, ceosalary$age)
y <- ceosalary$salary
beta <- solve(t(X) %*% X) %*% t(X) %*% y
y_hat <- X %*% beta

# R2 
SSR <- sum((y - y_hat)^2)
SST <- sum((y - mean(y))^2)
R2 <- 1 - (SSR / SST)

1 - var(m1$residuals)/var(m1$model$salary)

# MSE
sum(m1$residuals^2) / m1$df.residual

# adj. R2
1 - (sum(m1$residuals^2) / m1$df.residual) / var(m1$model$salary)
```
