---
title: "Lab 2"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

# Data

In this lab, we will continue to work with data on CEO salaries from 1990, from the dataset `ceosalary.csv`.

The relevant variables are:

-   `salary`: 1990 compensation (\$1000s)

-   `age`: Age (years)

-   `profits`: 1990 profits (\$ millions)

-   `grad`: binary for graduate school

```{r}
ceosalary <- read.csv("data/ceosalary.csv")
```

# Model Fit

$$salary_i = \beta_0 + \beta_1 profits_i + u_i$$

```{r}
m1 <- lm(salary ~ profits, ceosalary)
summary(m1)
```

## R Squared

$$R^2 = 1 - \frac{SSR}{SST} = 1 - \frac{\Sigma_N (y_i - \hat{y_i})^2}{\Sigma_N (y_i - \bar{y})^2}$$ $R^2$ is the proportionate reduction in error associated with the dependent variable when moving from an intercept-only model to the model of interest. It is the proportion of total variance in the dependent variable that is accounted for by the model.

```{r}
# observed y 
y <- ceosalary$salary

# predicted y 
X <- cbind(1, ceosalary$profits)
beta <- solve(t(X) %*% X) %*% t(X) %*% y
y_hat <- X %*% beta

# sum of squared residuals
SSR <- sum((y - y_hat)^2)

# total sum of squares
SST <- sum((y - mean(y))^2)

# r squared
1 - (SSR / SST)
1 - var(m1$residuals)/var(m1$model$salary)
```

## Mean Squared Error

$$MSE = \frac{\Sigma_N (y_i - \hat{y_i})^2}{N - K - 1} $$ Mean squared error is the estimated variance of prediction errors, or the estimated residual variance, $\hat{\sigma}^2$.

```{r}
# mse
SSR / (177 - 1 - 1) 
sum(m1$residuals^2) / m1$df.residual
```

## Adjusted R Squared

$$R_{adj}^2 = 1 - \frac{MSE}{Var(y)} = 1 - \frac{\frac{\Sigma_N (y_i - \hat{y_i})^2}{N - K - 1}}{\frac{\Sigma_N (y_i - \bar{y})^2}{N - 1}}$$

Adjusted $R^2$ is mean squared error divided by the variance of the dependent variable. The difference between adjusted and unadjusted $R^2$ is that the former divided by the residual degrees of freedom (sample size minus parameters estimated), and thus adjusts $R^2$ downwards to account for the increase in fit associated with adding additional parameters, regardless of how well they predict the dependent variable out-of-sample.

```{r}
1 - (sum(m1$residuals^2) / m1$df.residual) / var(m1$model$salary)
```

# Multiple Regression

$$salary_i = \beta_0 + \beta_1 profits_{i} + \beta_2 age_{i} + u_i$$

```{r}
m2 <- lm(salary ~ profits + age, ceosalary)
summary(m2)
```

## Comparing Model Fit

How does $\hat{\beta}_1$ change from `m1` to `m2`?

How does $R^2$ change from `m1` to `m2`?

How does adjusted $R^2$ change from `m1` to `m2`?

## Plotting Multiple Regression

Plot the relationship between profits and salary. Be careful! Consider the interpretation of $\hat{\beta}_1$.

```{r, include=FALSE}
# Load libraries
library(ggplot2)
library(dplyr)
```

```{r, warning=FALSE}
# Fix age at its mean and vary profits for the profits-salary relationship
profits_range <- data.frame(
  profits = seq(min(ceosalary$profits), max(ceosalary$profits), length.out = 100),
  age = mean(ceosalary$age)
)
profits_range$salary <- predict(m2, newdata = profits_range)

# base R 
plot(profits_range$profits, profits_range$salary, type = "l",
     xlab="Firm Profits (million USD)", ylab="CEO Salary (thousand USD)",
     main = "CEO Salary by Firm Profits at Mean Age")

# ggplot
ggplot(profits_range, aes(x = profits, y = salary)) +
  geom_line(size = 1) +
  labs(
    title = "Relationship Between Profits and Salary",
    x = "Profits",
    y = "Salary") +
  theme_minimal()
```

Add the relationship between age and salary to the same plot.

```{r, eval=FALSE}
# Fix profits at its mean and vary age for the age-salary relationship
age_range <- data.frame(
  age = seq(min(ceosalary$age), max(ceosalary$age), length.out = 100),
  profits = mean(ceosalary$profits)
)
age_range$salary <- predict(m2, newdata = age_range)

# combine datasets
profits_range$variable <- "Profits"
age_range$variable <- "Age"
combined_data <- bind_rows(profits_range, age_range)

ggplot() +
  geom_line(data = age_range, aes(x = age, y = salary), color = "red") +
  geom_line(data = profits_range, aes(x = profits, y = salary), color = "blue") +
  scale_x_continuous(name = "Age (Years)", 
                     sec.axis = sec_axis(~ ., name = "Profits (Million USD)")) +
  labs(
    y = "Salary (Thousand USD)",
    title = "CEO Salary by Age and Firm Profits"
  ) +
  theme_minimal() +
  theme(
    axis.title.x.bottom = element_text(color = "red"),
    axis.title.x.top = element_text(color = "blue"),
    axis.line = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5)
  )
```

```{r, echo=FALSE}
# Fix profits at its mean and vary age for the age-salary relationship
age_range <- data.frame(
  age = seq(min(ceosalary$age), max(ceosalary$age), length.out = 100),
  profits = mean(ceosalary$profits)
)
age_range$salary <- predict(m2, newdata = age_range)

# normalize the x-axis values 
profits_range$normalized_x <- scale(profits_range$profits, center = TRUE, scale = TRUE)
age_range$normalized_x <- scale(age_range$age, center = TRUE, scale = TRUE)

# combine datasets
profits_range$variable <- "Profits"
age_range$variable <- "Age"
combined_data <- bind_rows(profits_range, age_range)

# define transformations for the axes
profits_trans <- function(x) x * sd(profits_range$profits) + mean(profits_range$profits)
age_trans <- function(x) x * sd(age_range$age) + mean(age_range$age)

# plot
ggplot(combined_data, aes(x = normalized_x, y = salary, color = variable)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Profits" = "blue", "Age" = "red"), name = "Variable") +
  scale_x_continuous(
    name = "Age (years)",  # Bottom axis
    breaks = seq(-2, 2, 0.5),
    labels = round(age_trans(seq(-2, 2, 0.5)), 0),  # Convert back to Age scale
    sec.axis = sec_axis(
      ~ .,
      name = "Profits (million USD)",  # Top axis
      breaks = seq(-2, 2, 0.5),
      labels = round(profits_trans(seq(-2, 2, 0.5)), 0)  # Convert back to Profits scale
    )
  ) +
  labs(
    title = "Relationship Between Profits, Age, and Salary",
    y = "Salary (thousand USD)"
  ) +
  theme_minimal() +
  theme(
    axis.title.x.bottom = element_text(color = "red"),
    axis.title.x.top = element_text(color = "blue"),
    axis.text.x.bottom = element_text(color = "red"),
    axis.text.x.top = element_text(color = "blue"),
    panel.grid.minor = element_blank()
  )

```
