---
title: "Lab 5"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Logs and Exponents

## Inverse Functions

$$ln(e^x) = x$$ $$e^{ln(x)} = x$$ Note: the `log()` function computes
natural logs by default. The `exp()` function computes natural
exponential functions.

```{r}
x <- 10
log(exp(x)) 
exp(log(x))
```

## Helpful Rules

$$ exp(a + b) = e^{a + b} = e^ae^b $$
$$ exp(a - b) = e^{a - b} = \frac{e^a}{e^b} $$
$$ ln(a) + ln(b) = ln(ab)$$ $$ ln(a) - ln(b) = ln(\frac{a}{b})$$
$$\frac{d}{dx} ln(x) = \frac{1}{x}$$

```{r}
a <- 10; b <- 7 

exp(a + b) == exp(a) * exp(b)

round(exp(a - b)) == round(exp(a) / exp(b)) 

log(a) + log(b) == log(a*b)

round(log(a) - log(b)) == round(log(a/b))

D(expression(log(x)), "x")
```

## 1-Unit Change

A 1-unit increase in ln(x) implies a non-constant effect in x. Constant
increases in ln(x) are associated with *order of magnitude* increases in
x.

```{r}
log(10*x) - log(x) 
log(100*x) - log(10*x)
log(1000*x) - log(100*x)

# log(a) - log(b) == log(a/b)
log((10*x)/x)
log(10)
```

# Log Transformations

In this lab, we will return to the CEO salaries data set.

-   `salary`: 1990 compensation (\$1000s)

-   `profits`: 1990 profits (\$ millions)

```{r}
ceosalary <- read.csv("data/ceosalary.csv")
```

Create a variable `lsalary` that takes the log of salary in dollars.
Create a variable `lprofits` that takes the log of profits in dollars.

What type of data will be lost in this transformation?

```{r}
ceosalary$lsalary <- log(ceosalary$salary * 1000)
ceosalary$lprofits <- log(ceosalary$profits * 1000000)
```

# Interpreting Logs

## Level-Level Model

1 unit change in $x_1$ is associated with a $\beta_1$ unit change in y.
Remember: units of x (millions) and units of y (thousands)!

```{r}
m1 <- lm(salary ~ profits, ceosalary)
summary(m1)
```

## Log-Level Model

$$ ln(y_i) = \beta_0 + \beta_1 x_{1i} + u_i$$

Factor Change: 1 unit change in $x_1$ multiplies $E(y)$ by
$e^{\beta_1}$. Percent Change: 1 unit change in $x_1$ is associated with
a $100(e^{\beta_1(\Delta x_1)} - 1)$ *percent* change in $E(y)$.

```{r}
m2 <- lm(lsalary ~ profits, ceosalary)
summary(m2)

# factor change 
exp(coef(m2)[2])

# percent change
100 * exp(coef(m2)[2] - 1)
```

## Level-Log Model

$$ y_i = \beta_0 + \beta_1 ln(x_{1i}) + u_i$$

Unit Change: A 1 *percent* increase in $x_1$ is associated with a
$0.01 \beta_1$ unit increase in $E(y)$.

$$ y_i = \beta_0 + \beta_1 ln(x_{1i}) + u_i$$

```{r}
m3 <- lm(salary ~ lprofits, ceosalary)
summary(m3)

# unit change 
0.01 * coef(m3)[2]
```

## Log-Log Model

$$ ln(y_i) = \beta_0 + \beta_1 ln(x_{1i}) + u_i$$

Percent Change: 1 *percent* change in $x_1$ is associated with a
$\beta_1$ *percent* change in $E(y)$.

```{r}
m4 <- lm(lsalary ~ lprofits, ceosalary)
summary(m4)

# percent change 
coef(m4)[2]
```

# Hypothesis Testing

$$ ln(salary_i) = \beta_0 + \beta_1 ln(profits_i) + u_i$$
$$ H_0: \beta_1 = 0 $$

```{r}
# list-wise delete
data <- na.omit(ceosalary[, c("lsalary", "lprofits")])

# construct X
X <- cbind(rep(1, nrow(data)), data[, "lprofits"])

# construct y
y <- data[, "lsalary"]

# df 
df <- nrow(X) - ncol(X)

# OLS
B <- solve(t(X)%*%X) %*% t(X)%*%y

# residuals
res <- y - X%*%B

# estimate sig^2
sig2 <- sum(res^2) / df

# estimate var-cov matrix for B
V <- sig2 * solve(t(X)%*%X)

# ses
se <- sqrt(diag(V))

# t stats
t <- B / se

# p
p <- 2*pt(t, df, lower.tail = FALSE)

# confidence intervals 
lci <- B - qt(0.975, df)*se
uci <- B + qt(0.975, df)*se

# table
knitr::kable(round(data.frame(beta=B, se=se, tstat=t, p=p, lci=lci, uci=uci, row.names = c("Intercept","Logged Profits")), 3))
```

How do we interpret the values in the table?

## F test

Compares the residual SS of the restricted and unrestricted models.

$$
F = \frac{SSR_r - SSR_{ur} / q}{SSR_{ur} / (N-K-1)}
$$

Where q is the number of restrictions. The F-statistic measures the
ratio of explained variance to unexplained variance. A high F-statistic
suggests that the additional parameter(s) significantly improve the
model.

```{r}
# unrestricted
m5 <- lm(lsalary ~ lprofits + lsales, ceosalary)

# re-estimate restricted model on unrestricted data
m5_r <- lm(lsalary ~ lprofits, data = m5$model)
m5_r <- update(m5, . ~ . - lsales, data = m5$model)

# SSR, for restricted and unrestricted models
ssr_r <- sum(m5_r$residuals^2)
ssr_ur <- sum(m5$residuals^2)

# df unrestricted
df_ur <- m5$df.residual

# df difference, # restrictions
q <- m5_r$df.residual - m5$df.residual

# F stat
Fstat <- ((ssr_r - ssr_ur) / q) / (ssr_ur / df_ur)

# p value
pf(Fstat, q, df_ur, lower.tail=F)

# compare to anova
anova(m5_r, m5)
```
