---
title: "Lab 6"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

# Log Interpretation  

In this lab, we will return to the CEO salaries data set. 

  - `salary`: 1990 compensation ($1000s)
  - `lsalary`: 1990 compensation ($1000s) logged 
  - `profits`: 1990 profits ($ millions)
  - `lprofits`: 1990 profits ($ millions) logged 
  
```{r}
ceosalary <- read.csv("data/ceosalary.csv")

ceosalary$lsalary <- log(ceosalary$salary)
ceosalary$lprofits <- log(ceosalary$profits)
```

## Log-Level Model 

$$ ln(y_i) = \beta_0 + \beta_1 x_{1i} + u_i$$
Precise: 1 unit change in $x_1$ is associated with a $100(e^{\beta_1\Delta x_1} - 1)$ *percent* change in $E(y)$.

Approx: 1 unit change in $x_1$ is associated with a $100\beta_1$ *percent* change in $E(y)$.

```{r}
m2 <- lm(lsalary ~ profits, ceosalary)
summary(m2)

# precise
100 * (exp(coef(m2)[2]) - 1)

# approx
100 * coef(m2)[2] 
```

## Level-Log Model 

$$ y_i = \beta_0 + \beta_1 ln(x_{1i}) + u_i$$
Precise: p *percent* change in $x_1$ is associated with a $\beta_1 ln(\frac{100 + p}{100})$ unit change in $E(y)$.

Approx: 1 *percent* change in $x_1$ is associated with a $\frac{\beta_1}{100}$ unit change in $E(y)$.

```{r}
m3 <- lm(salary ~ lprofits, ceosalary)
summary(m3)

# precise 
coef(m3)[2] * log(1.01)

# approximation
coef(m3)[2] / 100
```

## Log-Log Model 

$$ ln(y_i) = \beta_0 + \beta_1 ln(x_{1i}) + u_i$$
Precise: p *percent* change in $x_1$ is associated with a $100(e^{\beta_1 ln(\frac{100+p}{100})} - 1)$ *percent* change in $E(y)$.

Approx: 1 *percent* change in $x_1$ is associated with a $\beta_1$ *percent* change in $E(y)$.  

```{r}
m4 <- lm(lsalary ~ lprofits, ceosalary)
summary(m4)

# precise 
100*(exp(coef(m4)[2]*log(101/100)) - 1)

# approx 
coef(m4)[2]
```

# Conditional Marginal Effects 

When the effect is polynomial or there is an interaction with another independent variable, the marginal effect is conditional. 

To find the conditional marginal effect, we take the first partial derivative with respect to $x_1$. 

## Quadratics 

Let's try a simple example with a quadratic using the `trees` data.

$$volume_i = \beta_0 + \beta_1height_i + \beta_2(height_i)^2 + u_i$$
Conditional Marginal Effect of Height on Volume: 
$$
\frac{\partial volume_i}{\partial height_i} = \beta_1 + 2\beta_2height_i
$$ 
```{r echo=TRUE}
# load data
trees <- datasets::trees

# estimate quadratic model
m <- lm(Volume ~ Height + I(Height^2), data=trees)

# effect = B1 + 2B2(height) 

# range of height
height <- seq(min(trees$Height), max(trees$Height), 0.1)

# calculate conditional marginal effect
cme_height <- coef(m)["Height"] +  2 * height * coef(m)["I(Height^2)"]

# plot cme
plot(height, cme_height, 
     main = "Conditional Marginal Effect of Height Across Values of Height",
     ylab="CME of Height", xlab="Height", type="l",
     xlim=c(60,90), ylim=c(-2, 4))
abline(h=0, lty=3)
```

Predicted Volume by Height:  

$$volume_i = \beta_0 + \beta_1height_i + \beta_2(height_i)^2 + u_i$$
```{r}
height <- data.frame(Height = height)
volume <- predict(m, newdata = height)

plot(height$Height, volume, 
     main = "Predicted Volume by Height",
     ylab="Volume", xlab="Height", type="l",
     xlim=c(60,90), ylim=c(10, 60))
```

## Interactions

$$
volume_i = \beta_0 + \beta_1 girth_i + \beta_2 height_i + \beta_3 girth_i * height_i
$$

Predicted Volume by Height at Different Girths: 

- This is the same code you've been practicing since PS1. 
- Will the lines be parallel? 

```{r}
# estimate interaction model
mi <- lm(Volume ~ Height*Girth, data=trees)

# range of height values 
height <- seq(min(trees$Height), max(trees$Height), 0.1)

# girth values 
girth <- c(10, 15, 20)

new_data <- expand.grid(Height = height, Girth = girth)

#predict volume based on height and girth in new dataset 
predicted_volume <- predict(mi, newdata = new_data)

#combine new dataset with predicted volume 
predicted_df <- cbind(new_data, Volume = predicted_volume)

# install.packages("ggplot2")
library(ggplot2)

# ggplot 
ggplot(predicted_df, aes(x = Height, y = Volume, color = factor(Girth))) +
  geom_line() +
  scale_color_manual(values = c("blue", "red", "green"),
                     labels = c(paste("Girth = 10"), 
                                paste("Girth = 15"), 
                                paste("Girth = 20")),
                     name = "Tree Girth") +
  labs(x = "Tree Height", y = "Predicted Tree Volume",
       title = "Predicted Volume vs. Height at Different Girth") +
  theme_minimal()
```

