---
title: "Lab 8"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r}
#library(haven)
#data <- read_dta("data.dta")
#write.csv(data,"~/path/data.csv", row.names = FALSE)
raw_resume <- read.csv("data/resume.csv")
```

# Data 

Resume data is from an experiment that manipulated attributes in fictitious resumes, which they sent out to employers, and measured whether the resumes received a call. 

- `call` is the dependent variable of interest (did the employer call the fictitious applicant for an interview) (binary) 
- `black` is the treatment variable in the data set (whether the resume has a "Black-sounding" name) (binary) 
- `yearsexp` years of experience (integer; 1 - 44)
- `female`gender (binary) 
- `computerskills` computer skills (binary) 
- `ofjobs` number of previous jobs (integer; 1 - 7)
- `education` applicants education level (categorical; 0-4) 
  - 0: Education not reported
  - 1: High school dropout
  - 2: High school graduate
  - 3: Some college
  - 4: College graduate or higher
  
Bertrand, M., & Mullainathan, S. (2004). Are Emily and Greg more employable than Lakisha and Jamal? A field experiment on labor market discrimination. American economic review, 94(4), 991-1013

```{r, warning=F}
#install.packages("fastDummies")
library(fastDummies)

resume <- dummy_cols(raw_resume, select_columns = 'education')
```

# Categorical Variables 

Estimate a model, using OLS, that regresses `call` on `black`, `yearsexp`, `education`, and `female`.
```{r}
m1 <- lm(call ~ black + female + factor(education) + yearsexp, raw_resume)

summary(m1)

m1 <- lm(call ~ black + female + education_1 + education_2 + education_3 + education_4 + yearsexp, resume)

summary(m1)
```

# F Tests

Compares the residual SS of the restricted and unrestricted models.

$$
F = \frac{SSR_r - SSR_{ur} / q}{SSR_{ur} / (N-K-1)}
$$
Where q is the number of restrictions. The F-statistic measures the ratio of explained variance to unexplained variance.A high F-statistic suggests that the additional parameter(s) significantly improve the model.

Conduct an F-test to see if `ofjobs` has an effect on the dependent variable

```{r}
# unrestricted 
ur <- lm(call ~ black + female + education_1 + education_2 + education_3 + education_4 + yearsexp + ofjobs, resume)

# estimate restricted model on unrestricted data
r <- update(ur, . ~ . - ofjobs, data = ur$model)

# SSR, for restricted and unrestricted models
ssr_r <- sum(r$residuals^2)
ssr_ur <- sum(ur$residuals^2)

# df unrestricted
df_ur <- ur$df.residual

# df difference, # restrictions
q <- r$df.residual - ur$df.residual

# F stat
Fstat <- ((ssr_r - ssr_ur) / q) / (ssr_ur / df_ur)

# p value
pf(Fstat, q, df_ur, lower.tail=F)

# compare to anova
stats::anova(r, ur)
```

Conduct a hypothesis test to test the null hypothesis that the effect of having a 'Black-sounding' name, compared to having a 'White-sounding' name, is equal to the effect of having a 'female-sounding' name, compared to having a 'male-sounding' name. 

```{r, warning=F}
library(car)
car::linearHypothesis(m1, "black - female = 0")
```

# Bootstrapping 

Calculate bootstrapped 95% confidence intervals for `black`. Use the standard bootstrap approach of taking random draws, with replacement, from the rows of your model’s data matrix. Use the percentile method to calculate the confidence bounds. 

```{r}
# number of bootstrap samples
boots <- 1000

# number of regression coefficients (K+1)
#ncoef <- summary(m1)$df[1] 
ncoef <- 1

# initialize array to store coefs
boots_beta <- array(NA, c(boots, ncoef))

# iterate over samples
for (i in 1:boots){
  
  # sample rows of data w/ replacement
  new <- sample(1:nrow(m1$model), nrow(m1$model), replace = T) 
  
  # run model with bootstrap sample
  boots_beta[i, ] <- coef(lm(formula(m1), data = m1$model[new, ]))[2] 
}

# calculate SE for `black`
boots_se <- apply(boots_beta, 2, sd) 
sd(boots_beta)

# calculate CI 
lci <- coef(m1)[2] - qt(0.975, m1$df.residual)*boots_se
uci <- coef(m1)[2] + qt(0.975, m1$df.residual)*boots_se

print (c(lci, uci))
```

Now calculate confidence intervals with the BCa method. 

```{r, warning=F}
# boot package 
library(boot)

boot_fn <- function(data, indices) {
  d <- data[indices, ]  # subset data by randomly generated vector of row indices (w replacement)
  fit <- update(m1, . ~ ., data = d)
  B <- coef(fit)[2]
  se <- sqrt(diag(vcov(fit)))[2] 
  return(c(B, se))
}

# boot call
m_boot <- boot(data = m1$model, statistic = boot_fn, R = 10000)

# use boot.ci to get CIs
boot.ci(m_boot, type = "bca")
```

# Interactions 

Estimate a new model, using OLS, that regresses `call` on `black`, `yearsexp`, and the interaction between them. 

```{r}
m2 <- lm(call ~ black*yearsexp, resume)
summary(m2)
```

# Marginal Effects 

Make a plot that shows how the expected value of the dependent variable changes across years of experience (`yearsexp`), with separate lines for the two race treatment conditions (`black`). 
```{r}
# range of years experience  
range_yearsexp <- seq(min(resume$yearsexp), max(resume$yearsexp), 1)

# race values 
values_black <- c(0,1)

new_data <- expand.grid(yearsexp = range_yearsexp, black = values_black)

#predict call back based on years experience and race
predicted_call <- predict(m2, newdata = new_data)

#combine new dataset with predicted volume 
predicted_df <- cbind(new_data, call = predicted_call)

# install.packages("ggplot2")
library(ggplot2)

# ggplot 
ggplot(predicted_df, aes(x = yearsexp, y = predicted_call, color = factor(black))) +
  geom_line() +
  scale_color_manual(values = c("blue", "red"),
                     labels = c(paste("White"), 
                                paste("Black")),
                     name = "Race") +
  labs(x = "Years Experience", y = "Predicted Call Back",
       title = "Predicted Call vs. Years Experience") +
  theme_minimal()
```

Use the marginaleffects package and the delta method to calculate 95% confidence bounds for the conditional marginal effect of `black` at each value of `yearsexp`. 

```{r}
library(marginaleffects)
options(marginaleffects_print_omit = c("s.value"))
?marginaleffects::slopes # s.value = Shannon information

# estimate ME of black at range of yearsexp
cme <- slopes(m2, variables = c("black"),
       newdata = datagrid(yearsexp = range_yearsexp))

# plot ME and SE for range of yearsexp
plot(cme$yearsexp, cme$estimate,
     main = "CME of Black Across Years Experience",
     ylab="CME of Black", xlab="Years Experience",
     type = "l", ylim=c(-0.2,0.2), xlim=c(0,45))
abline(h=0, lty=3)
lines(cme$yearsexp, cme$estimate + 1.96*cme$std.error) 
lines(cme$yearsexp, cme$estimate - 1.96*cme$std.error) 
```

# Chow Test 

The Chow test is a special form of F-test used to test the null hypothesis that a linear model is identical across groups

$$
F = \frac{SSR_{pooled} - (SSR_{male} + SSR_{female})}{SSR_{pooled}} \cdot \frac{N - 2(K+1)}{K+1}
$$

Estimate the same model but for male and female respondents. Conduct a Chow test of the null hypothesis that the two groups (for each category of gender) can be combined into a single model (i.e., that the true model is the same across these two groups).

```{r}
# male <- subset(resume, female == 0)
male <- resume[resume$female == 0, ]

# female <- subset(resume, female == 1)
female <- resume[resume$female == 1, ]

# models (pooled, male, female) 
m2p <- lm(call ~ black * yearsexp, data = resume)
m2m <- lm(call ~ black * yearsexp, data = male)
m2f <- lm(call ~ black * yearsexp, data = female)

ssr_p <- sum(m2p$residuals^2)
ssr_m <- sum(m2m$residuals^2)
ssr_f <- sum(m2f$residuals^2)

n <- nrow(m2p$model)
k <- as.numeric(length(m2p$coefficients))

# chow
chow <- ( (ssr_p - (ssr_m + ssr_f)) / ssr_p ) * ( (n - 2*(k+1)) / (k+1) )

# p value
pf(chow, df1 = k, df2 = (n - (2 * k)), lower.tail =F)
```

