---
title: "Lab 8"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

# PS7 Problem 2

We will use the following variables in the dataset `contribupdate.csv`:

-   `age`: Age (years)
-   `female`: 1=female, 0=male
-   `faminc`: Household income (in \$1,000 U.S. dollars);
-   `given`: Campaign contributions over the past year (in U.S. dollars).

```{r}
contrib <- read.csv("data/contribupdate.csv")
contrib$ln_faminc <- log(contrib$faminc) 
contrib$ln_given <- log(contrib$given)
```

$$
 \ln(\text{given}_i) = \beta_0 + \beta_1 \text{age}_i + \beta_2 \text{female}_i + \beta_3 \ln(\text{faminc}_i) + u_i
$$

```{r}
m2 <- lm(ln_given ~ age + female + ln_faminc, contrib)
```

Calculate the following quantities by hand and use simulation to get 95% confidence bounds. 

```{r}
library(arm)
sims <- arm::sim(m2, n.sims = 1000)
beta <- as.matrix(sims@coef)
```

## The predicted *percent* change in `given` for a 10-year increase in `age`. 

Log-Level Model: Unit $\Delta x_1$ is associated with a $100(e^{\beta_1\Delta x_1} - 1)$ *percent* change in $E(y)$.

```{r, echo=T}
# by hand 
(exp(10*coef(m2)["age"]) - 1)*100

# simulations
pc_sims <- (exp(10 * beta[, "age"]) - 1) * 100

# confidence interval 
quantile(pc_sims, c(0.025, 0.975))
```

The predicted change in campaign contributions (`given`) for a 10-year increase in age (`age`) is about 12%, holding all else constant. The 95% confidence interval is between 10.5% and 13.8%. 

## The difference in the expected values of `given` for females versus males, holding `age` and `faminc` at their respective means.

```{r, echo=T}
# BY HAND

# ln(given) for females 
m2f <- coef(m2)["(Intercept)"] + coef(m2)["age"] * mean(contrib$age) + coef(m2)["female"] * 1 + coef(m2)["ln_faminc"] * mean(contrib$ln_faminc)

# ln(given) for males 
m2m <- coef(m2)["(Intercept)"] + coef(m2)["age"] * mean(contrib$age) + coef(m2)["female"] * 0 + coef(m2)["ln_faminc"] * mean(contrib$ln_faminc)

# smearing factor 
norm <- exp(summary(m2)$sigma^2 / 2) # assuming normal 
duan <- mean(exp(residuals(m2))) # not assuming normal 

# assuming normal 
(exp(m2m) * norm) - (exp(m2f) * norm)

# not assuming normal 
(exp(m2m) * duan) - (exp(m2f) * duan)
```

```{r}
# SIMULATIONS

# ln(given) for females 
m2f_sims <- beta[,"(Intercept)"] + beta[,"age"] * mean(contrib$age) + beta[,"female"] * 1 + beta[,"ln_faminc"] * mean(contrib$ln_faminc)

# ln(given) for males 
m2m_sims <- beta[,"(Intercept)"] + beta[,"age"] * mean(contrib$age) + beta[,"female"] * 0 + beta[,"ln_faminc"] * mean(contrib$ln_faminc)

# difference in given bt males and females
diff_sims <- (exp(m2m_sims) * norm) - (exp(m2f) * norm)
diff_sims <- (exp(m2m_sims) * duan) - (exp(m2f) * duan)

# confidence interval 
quantile(diff_sims, c(0.025, 0.975))
```

The difference in the expected values of `given` for females versus males, holding `age` and `faminc` at their respective means is about \$61. In other words, men with average age and income contribute \$61 more to campaigns than women with average age and income. The 95% confidence interval is between \$53 and \$70.

## The average marginal effect of `faminc` on `given`.

Express in levels: 

$$given = e^{\beta_0 + \beta_1 \text{age}_i + \beta_2 \text{female}_i + \beta_3 \ln(\text{faminc}_i) + u_i}$$ 

Differentiate given with respect to faminc: 

$$\frac{\partial(given)}{\partial(faminc)} = \hat{given} \cdot \frac{\beta_3}{faminc}$$
    
```{r, echo=T}
# BY HAND (predict given, transform,  smear)

# assuming normal 
mean(coef(m2)["ln_faminc"] * (exp(predict(m2)) * norm / exp(contrib$ln_faminc))) # length? 

# not assuming normal (Duan)
mean(coef(m2)["ln_faminc"] * (exp(predict(m2)) * duan / exp(contrib$ln_faminc)))
```

```{r}
# SIMULATIONS 

me_sims <- array(NA, 1000)

for (i in 1:1000) {
  
  given_sims <- exp(beta[i,"(Intercept)"] + beta[i,"age"] * contrib$age + beta[i,"female"] * contrib$female + beta[i,"ln_faminc"] * contrib$ln_faminc) * duan # length?
  
  me_sims[i] <- mean(beta[i, "ln_faminc"] * (given_sims / exp(contrib$ln_faminc))) # length? 
  
}

# confidence interval
quantile(me_sims, c(0.025, 0.975))
```

The average marginal effect of `faminc` on `given` is about 2, which suggests that an \$1000 increase in household income is associated with \$2 increase in campaign contributions. The 95% confidence interval is between \$1.95 and \$2.19. 

# Heteroskedasticity 

We're going to return to the Chile data set. The variables of interest are as follows:

-   *vote_yes*: 1=will vote yes for Pinochet, 0=will vote no against Pinochet
-   *statusquo*: higher values = more support for the status quo, standardized to have mean zero and standard deviation of one
-   *income_1000*: monthly income, in thousands of pesos
-   *education*: factor variable with 3 levels: primary only (baseline), secondary (S), post-secondary (PS)
-   *sex*: factor variable with two levels, female (baseline) and male (M)

```{r}
# data
chile <- read.csv("data/chile.csv")
chile <- na.omit(chile)

# re-code income to 1000s
chile$income_1000 <- chile$income / 1000

# re-code education as dummies 
chile$educS <- ifelse(chile$education == "S", 1, 0)
chile$educPS <- ifelse(chile$education == "PS", 1, 0)

# re-code sex as dummy 
chile$female <- ifelse(chile$sex == "F", 1, 0)

# re-code interaction as product term 
chile$inc_sq <- chile$income_1000 * chile$statusquo
```

$$
voteyes_i = \beta_0 + \beta_1female_i + \beta_2educS_i + \beta_3educPS_i + \beta_4income1000_i + \beta_5statusquo_i + \beta_6(income1000_i)(statusquo_i) + u_i
$$

```{r}
m1 <- lm(vote_yes ~ female + educS + educPS + income_1000 + statusquo + inc_sq, data=chile)
summary(m1)
```

Heteroskedasticity: the variance of the error term is correlated with the independent variable(s). 

Error Term: the difference between the true value of $y$ and the expected value of $y$. 

In a linear probability model, the true value of $y$ takes on values of only 0 or 1, while the expected value of $y$ can be anything between 0 and 1.

The variance of the error term will vary depending on the level of predicted y, leading to heteroskedasticity.

## Robust (Huber-White) SE 

By Hand: substitute a diagonal matrix where the entries are each observation’s squared residual

$$Var(\hat{\beta}) = (X'X)^{-1}X' \Sigma X (X'X)^{-1}$$ 
Default: $\Sigma=\sigma^2I$

Robust: $\hat{\Sigma}=\hat{u}^2I$

```{r}
# YOUR CODE HERE
N <- nrow(m1$model)
K <- summary(m1)$df[1]

# add intercept column, remove dv column 
X <- cbind(rep(1,N), m1$model[,-1]) 
colnames(X)[1] <- "Intercept"
X <- as.matrix(X)

# default standard error 
se_default <- sqrt(diag((sum(m1$residuals^2) / m1$df.residual) * (solve(t(X) %*% X))))

# robust standard error (plug squared residuals in)
se_robust <- sqrt((diag(solve(t(X) %*% X) %*% (t(X) %*% diag(m1$residuals^2) %*% X) %*% solve(t(X) %*% X))))

# t-values 
t_values <- m1$coefficients / se_robust

# p-values 
p_values <- 2 * (1 - pt(abs(t_values), m1$df.residual))

# print  
round(data.frame(coefficient = coef(m1), 
           se.robust = se_robust, 
           t.value = t_values, 
           p.value = p_values), 5) 
```

Using sandwich/car

```{r}
library(sandwich)
library(car)

HC0_sw <- sqrt(diag(vcovHC(m1, type="HC0")))
HC0_car <- sqrt(diag(hccm(m1, type="hc0")))

HC3_sw <- sqrt(diag(vcovHC(m1))) # default is HC3
HC3_car <- sqrt(diag(hccm(m1, type="hc3")))

round(data.frame(coefficient = coef(m1), 
                 default = se_default,
                 robust = se_robust, 
                 HC0 = HC0_sw,
                 HC3 = HC3_sw), 5) # round to 5 decimal places
```

In marginaleffects

```{r}
library(marginaleffects)
options(marginaleffects_print_omit = c("s.value"))

# new model with interaction term 
m1me <- lm(vote_yes ~ female + educS + educPS + income_1000*statusquo, data=chile)

# estimate ME of status quo at range of income

cme <- slopes(m1me, variables = c("statusquo"), newdata = datagrid(income_1000 = seq(min(chile$income_1000), max(chile$income_1000), 1))) # default SE 

cme_HC3 <- slopes(m1me, variables = c("statusquo"), newdata = datagrid(income_1000 = seq(min(chile$income_1000), max(chile$income_1000), 1)), vcov = "HC3") # robust SE 

# plot ME and SE for range of income
plot(cme$income_1000, cme$estimate,
     main = "CME of Status Quo Across Income",
     ylab="CME of SQ", xlab="Income (1000 Pesos)",
     type = "l", ylim=c(0.35,0.5), xlim=c(0,200))

# add default CI 
lines(cme$income_1000, cme$estimate + 1.96*cme$std.error, lty=2) 
lines(cme$income_1000, cme$estimate - 1.96*cme$std.error, lty=2) 

# add robust CI 
lines(cme$income_1000, cme$estimate + 1.96*cme_HC3$std.error, lty=2,col="red") 
lines(cme$income_1000, cme$estimate - 1.96*cme_HC3$std.error, lty=2,col="red")
```

## Diagnosing Heteroskedasticity 

F-Test (use the squared residuals as an estimate of error variance and check the extent to which they can be explained by IV)

```{r}
# regress squared residuals on predictors
m1_res_1 <- lm(m1$residuals^2 ~ m1$model[,2] + m1$model[,3] + m1$model[,4] + m1$model[,5] + m1$model[,6] + m1$model[,7])

# F-stat: tests whether at least one IV significantly explains variation in the squared residuals (heteroskedasticity)
# p-value: the probability of obtaining the observed F-statistic if the null hypothesis (homoskedasticity) were true
summary(m1_res_1) 
```

Breusch-Pagen Test (regress squared residuals on IV)

```{r}
# B-P test, use chi-squared test chi2 = N*R2, on K df
N <- sum(summary(m1_res_1)$df[1:2]) # num observations
R2 <- summary(m1_res_1)$r.squared # explained variance 
chi2 = N*R2 # test statistic 
K = summary(m1_res_1)$df[1] - 1 # num parameters 

1 - pchisq(q = chi2, df = K) # p-value 

#install.packages("lmtest")
library(lmtest)
bptest(m1)
```

# Fixed Effects 

Add group as categorical variable. Since we have removed what is shared across observations within groups, the errors are now independent, and SEs are unbiased

Include fixed effects for `region`. 

```{r}
m1_fe <- lm(vote_yes ~ female + educS + educPS + income_1000 + statusquo + inc_sq + as.factor(region), data=chile)
summary(m1_fe)
```

## Clustered Standard Errors 

```{r}
# with sandwich 
sqrt(diag(sandwich::vcovCL(m1, cluster = ~ region, type = "HC1")))
sqrt(diag(sandwich::vcovCL(m1, cluster = ~ region, type = "HC3")))
```