---
title: "mensik_ps12"
output: html_document
date: "2026-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

devtools::install_github("synth-inference/synthdid")
library(synthdid)
library(did)
library(fixest)
library(plm)
library(broom)
library(lmtest)
library(sandwich)
```

# 1

```{r}
data(california_prop99)
?california_prop99
```

## 1a.

```{r}
#head(california_prop99)
#class(california_prop99$Year)

m_fe <- feols(PacksPerCapita ~ treated | State + as.factor(Year), data = california_prop99, cluster = ~State
)


summary(m_fe)

```

## 1b. Interpret the coefficient on treated:

After California's cigarette tax was implemented in 1989, the difference between California's per-capita cigarette consumption and other states' decreased by 27.35 percentage points, holding all else equal.

## 1c. Calculate clustered standard errors by State

(HINT: the packages have functions for this available) and construct a 95% confidence interval for the effect of the CA tax hike on cigarette consumption.

```{r}
confint(m_fe, level = 0.95)

tidy(m_fe, conf.int = TRUE, conf.level = 0.95)

```

## 1d. estimate same model using lm

and show you get the same estimate for treated.

$Y_{it} = \alpha_i + \lambda_t + \beta(treated_{it}) + \epsilon_{it}$

treated = state x pre/post

```{r}
year_f <- as.factor(california_prop99$Year)

m_lm <- lm(PacksPerCapita ~ State + year_f + treated, data = california_prop99 )

summary(m_lm)

#coeftest(m_lm, vcovCL(m_lm, cluster =~State))

```

#2

## Problem 2

For this analysis, you will use the dataset `mpdta` from the package `did`. Load the data using `data(mpdta)`. Use `?mpdta` to see the data description and the variables available.

```{r}
?mpdta
```

a.  You first need to recode the `treat` variable so that it is 0 in years before treatment. Use this code below:

```{r echo=TRUE}
mpdta$treat_01 <- ifelse(mpdta$first.treat != 0 & mpdta$year >= mpdta$first.treat, 1, 0)
```

b.  Use `feols` in the `fixest` package to estimate a two-way fixed effects model to test the hypothesis that raising the minimum wage reduces teen employment. Use clustered standard errors by county.

first.treat = year that the state where the county is located raised its minimum wage, set to 0 for counties that have min wages equal to the fed min wage over the entire period

treat = whether/not a particular county is treated in that year

lpop = log of 1000s of population for the county

countyreal = unique ID for a particular county

year = year of the observation

$lemp_{it} = \beta(treat_{it}) + \alpha_i + \lambda_t + \epsilon_{it}$

```{r}

m1 <- feols(lemp ~ treat_01 | countyreal + year, data = mpdta, cluster =~countyreal)

summary(m1)
```

c.  Estimate a dynamic version of this model using the code below, then plot the model object using `iplot`. Is there evidence against the parallel trends assumption?

```{r echo=TRUE}
# estimate dynamic version of the model
m_dynamic  <- feols(lemp ~ sunab(first.treat, year) | countyreal + year,
                    data = mpdta, vcov = cluster ~ countyreal)

iplot(m_dynamic, xlab = "Years since minimum wage increase", ref.line = 0)
```

Yes - the treated countries have different different employment starting points before treatment (significant pre treatment coefficients).

d.  All evidence considered, would you say there is support for the claim that minimum wage hikes reduced youth employment?
