---
title: "mensik_ps12"
output: html_document
date: "2026-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

devtools::install_github("synth-inference/synthdid")
library(synthdid)
library(did)
library(fixest)
library(plm)


```

#1

```{r}
data(california_prop99)
?california_prop99
```


##1a. 
Using either the fixest or plm package, estimate a two-way fixed effects model to test the hypothesis that California’s cigarette tax hike in 1989 reduced per-capita cigarette consumption relative to the other control states. HINT: you need to explicitly call Year as a factor variable.

```{r}


```


##1b.
Interpret the coefficient on treated.

##1c.
Calculate clustered standard errors by State (HINT: the packages have functions for this available) and construct a 95% confidence interval for the effect of the CA tax hike on cigarette consumption.

##1d.
Now estimate the same model using lm, and show you get the same estimate for treated.


#2

```{r}

```

##2a. 

##2b.

##2c.

##2d.