---
title: "mensik_pset_10"
format: html
editor: visual
---

## Data

the `refug` dataset of judge decisions by the Canadian Federal Court of Appeal. The judges were making decisions about people seeking refugee status. The variables are as follows:

-   `decision`: the DV; 2=grants refugee's appeal, 1=denies appeal
-   decision_num is 1 = grants 0 = denies
-   `judge`: name of the judge hearing the case; a factor variable
    -   came in as character
    -   judge_factor is as factor
-   `nation`: nation of origin for the refugee making the appeal; a factor variable
-   `rater`: judgment of the merit of the appeal by an independent rater; a factor variable with two levels, yes or no
    -   rater_factor
-   `language`: language of the case, English or French; a factor variable
-   `location`: location of original claim by refugee, Montreal, Toronto, or other; a factor variable
-   `success`: the logit of the success rate ($\text{ln} \left( \frac{p}{1-p} \right)$) of all cases from refugee's nation of origin (numeric)

Load the data and name it refug.

```{r, echo=FALSE, warning=FALSE}
library(lmtest)
library(sandwich)
library(dplyr)

refug <- read.csv("/Users/kristinamensik/Documents/GitHub/regression-study/refug.csv")

# recode variable decision so that 1 equals grants the appeal and 0 equals denies the appeal (right now is 2 = grants the appeal and 1 = denies)
refug$decision_num <- ifelse(refug$decision == "yes", 1, 0)

# check judge -> recode to factor
refug$judge_factor <- factor(refug$judge)
levels(refug$judge_factor) #ordered alphabetically

#check rater -> factor
refug$rater_factor <- factor(refug$rater)
levels(refug$rater_factor)

#language to factor
refug$language_factor <- factor(refug$language)

#location to factor
refug$location <- factor(refug$location)

#success
unique(refug$success)
class(refug$success)

levels(refug$language_factor)
levels(refug$location)

```

## Problem 1

Estimate linear model using predictors excluding the 'nation' variable

```{r}

m1 <- lm(decision_num ~ judge_factor + rater_factor + language_factor + location + success, data = refug)

# clustered standard errors by nation

cov_clust_se <- vcovCL(m1, cluster = ~nation)

coeftest(m1, vcov. = cov_clust_se)
```

judgeHeald - the -0.238 coefficient means that having Judge Heald as opposed to Judge Desjardin is associated with a 0.24 decrease in the likelihood of having a case granted, holding all else constant. This effect is highly significant.

locationToronto - holding all else constant, having a case heard or seeking refugee status in Toronto instead of Montreal is associated with a 11.3 percentage point higher probability of being granted the case compared to a case/seeking refugee status in Montreal. However this is not statistically significant.

success - holding all else constant, a one unit increase in the logit of the success rate of all cases from the refugee's nation of origin is associated with a \~20.8 percentage point increase in the probability that a judge grants the appeal. A unit increase in logit is the natural log of the probability of a nation's success rate – so a country with a 30% success rate's logit would be ln(0.30/0.70) \~ -0.8. This is nonlinear, but roughly corresponds to fairly significant changes. It is statistically significant as well.

Coefficient plot - figure that plots the coef estimates for all independent variables along with their 95% confidence bounds
