---
title: "Mensik - PS10"
output:
  html_document:
    theme: united
    toc: true
    toc_float:
      collapsed: true
editor_options:
  chunk_output_type: console
date: "2025-08-18"
---

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}

library(ivreg)
library(ggplot2)
library(dplyr)
library(purrr)
library(car)
library(sandwich)
library(knitr)
library(kableExtra)
library(broom)
library(marginaleffects)
library(lmtest)
library(AER)
library(stringr)
library(tibble)


```

## Problem 11

### 1a. WLS: Resent on pslave1860

```{r, echo=FALSE}
#if laptop
abs_data <- read.csv("abs_data.csv")

#inspecting data

class(abs_data$water1860)

#class(abs_data$water1860)


#converting water to factor ?????

#convert state.abb to factor from chr
abs_data$state.abb_af <- factor(abs_data$state.abb)

#table(abs_data$state.abb_af)

#levels(abs_data$state.abb_af)

#natural log of coarea00
abs_data$ln_coarea00 <- log(abs_data$coarea00)

#summary(abs_data)

# latitude and longitude
abs_data$latitude_sq  = abs_data$latitude^2
abs_data$longitude_sq = abs_data$longitude^2
```

```{r, echo=FALSE}
#belatedly coming in to clean NAs -> put in new dataset

clean_abs <- c("resent", "pslave1860", "ln_coarea00" , "rugged" , "latitude"  , "longitude" , "latitude_sq" , "longitude_sq" , "water1860" , "state.abb_af", "sample.size", "cottonsuit")

cleaned_abs <- abs_data[complete.cases(abs_data[clean_abs]),]

```

```{r, echo=FALSE}
#wls model

#m1 <- lm(resent ~ pslave1860 + ln_coarea00 + rugged + latitude  + longitude + latitude_sq + longitude_sq + water1860 + state.abb_af, data = abs_data, weights = sample.size)

m01 <- lm(resent ~ pslave1860 + ln_coarea00 + rugged + latitude  + longitude + latitude_sq + longitude_sq + water1860 + state.abb_af, data = cleaned_abs, weights = sample.size)


#summary(m1)
#confint(m1, "state.abb_afWV", level = 0.95)

#summary(m01)

c_ols <- coeftest(m01, vcov. = vcovHC(m01, type = "HC3"))

#sum(is.na(abs_data$state.abb_afWV))

```

Note: I tested for heteroskedasticity on fitted model m01:

```{r eval=FALSE, include=FALSE}
#residuals vs. fitted plot (diagnostic, not formal -- useful for notation)
plot(m01$fitted.values, resid(m01),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals v Fitted")
abline(h=0, col = "red")


```

-   a BP Test showed evidence of heteroskedasticity

```{r eval=FALSE, include=FALSE}
bptest(m01)
```

-   as did White's test

```{r eval=FALSE, include=FALSE}

bptest(m01, ~ fitted(m01) + I(fitted(m01)^2), data = data.frame(fitted = fitted(m01))) 

```

Redid model interpretation with omitted NAs (should have done first) and reporting robust HC3 standard errors.

### 1b. Interpretation: pslave1860

Coefficient for pslave1860 is 0.1281. This means a one unit or 10% increase in the proportion of a county's 1860 population that was enslaved is associated with a 0.1281 percentage point increase in anti-Black attitudes today, holding all else constant.

However, there is no evidence we can reject the null -- the P is 0.45 over 0.05, the confidence interval crosses 0 (95% CI includes 0, p = 0.45).

Because of fixed effects for states, this reflects estimates of within-state comparisons.

Note on scale: -1 unit change in pslave1860 means +1 proportion, so +100percentage points (0 to 1) but that's nonseniscal, so a 1 percentage point increase (beta times 0.01) gives a 0.001281 percentage point on resentment.

```{r eval=FALSE, include=FALSE}
#confidence intervals by hand with robust SEs
B <- coef(m01)

V.robust_ols <- vcovHC(m01, type = "HC3")

se <- sqrt(diag(V.robust_ols)) 
  
cbind(
  B + qt(0.025, df=m01$df.residual) * se,
  B + qt(0.975, df = m01$df.residual) * se
)

```

### 1c. Interpretation: Water access

All else constant, counties with access to navigable water have an average of 0.052621 lower levels of racial resentment today than those without access, holding all else equal. This, however, is also not statistically significant (95% CI includes 0, p = 0.23)

### 1d. Interpretation: West Virginia

Alabama was excluded as the baseline, WV = 0.3845

All else constant, counties in West Virginia have 0.36845 higher levels of racial resentment today compared to those in Alabama - this would be surprising given the author's hypotheses since West Virginia had a smaller enslaved population. This is more significant than water and percent enslaved, but still short of statistical significance (95% CI crosses 0, p = 0.07).

## Problem 2

### 2a. ivreg

```{r include=FALSE}
m02 <- ivreg(resent ~  pslave1860 + 
               ln_coarea00 + rugged + latitude + longitude  + latitude_sq + longitude_sq + water1860 + state.abb_af 
             | cottonsuit + ln_coarea00 + rugged + latitude + longitude   + latitude_sq + longitude_sq + water1860 + state.abb_af, data = cleaned_abs, weights = sample.size)


c_2sls <- coeftest(m01, vcov. = vcovHC(m01, type = "HC3"))

#confint(m2, "pslave1860", level = 0.95)
```

```{r include=FALSE}
#confidence intervals by hand with robust SEs
B_2sls <- coef(m02)

V.robust_2sls <- vcovHC(m02, type = "HC3")

se_2sls <- sqrt(diag(V.robust_2sls)) 
  
cbind(
  B_2sls + qt(0.025, df=m02$df.residual) * se_2sls,
  B_2sls + qt(0.975, df = m02$df.residual) * se_2sls
)

```

### 2b. Table

```{r echo=FALSE, warning=FALSE}
#prep sig test vars

# pull pieces of robust models as named vectors
b_ols  <- c_ols[, "Estimate"]
se_ols <- c_ols[, "Std. Error"]
p_ols  <- c_ols[, "Pr(>|t|)"]

b_2sls  <- c_2sls[, "Estimate"]
se_2sls <- c_2sls[, "Std. Error"]
p_2sls  <- c_2sls[, "Pr(>|t|)"]

# star function (journal-style)
stars <- function(p) {
  ifelse(is.na(p), "",
    ifelse(p < .01, "***",
      ifelse(p < .05, "**",
        ifelse(p < .10, "*", "")
      )
    )
  )
}


#variables to display
vars <- c(
  "pslave1860",
  "ln_coarea00","rugged","latitude","longitude","latitude_sq","longitude_sq","water1860",
  grep("^state\\.abb_af", names(b_ols), value = TRUE)
)

#  classify rows for non "Sample" column
row_class <- function(v) {
  if (v == "pslave1860")          return("Main IV")
  if (str_detect(v, "^state\\.abb_af")) return("State FE")
  return("Control")
}

#constructing the table body
tbl_bod <- tibble::tibble(
  Variable = vars,
  Type = sapply(vars, row_class),
   OLS_Beta = b_ols[vars],
  OLS_SE   = se_ols[vars],
  OLS_p = p_ols[vars],
  `2SLS_Beta` = b_2sls[vars],
  `2SLS_SE`   = se_2sls[vars],
  `2SLS_p` = p_2sls[vars]
) %>%

#rounding
dplyr::mutate(
    OLS_Beta   = round(OLS_Beta, 3),
    OLS_SE     = round(OLS_SE,   3),
    `2SLS_Beta`= round(`2SLS_Beta`, 3),
    `2SLS_SE`  = round(`2SLS_SE`,   3)
  ) %>%

#stars to beta cols
  dplyr::mutate(
    OLS_Beta    = paste0(OLS_Beta, stars(OLS_p)),
    `2SLS_Beta` = paste0(`2SLS_Beta`, stars(`2SLS_p`))
  ) %>%
  # drop p cols
  dplyr::select(Variable, Type, OLS_Beta, OLS_SE, `2SLS_Beta`, `2SLS_SE`)


#observations (counties)
N_ols <- nobs(m01)
N_2sls <- nobs(m02)


#r2
r2_ols <- summary(m01)$r.squared
r2_2sls <- summary(m02)$r.squared

# sum of weights (respondents) if you want to show it
sum_w <- sum(model.weights(model.frame(m01)))

#footer rows to character for blank space
f_num <-function(x) ifelse(is.na(x), "", format(round(x,3), nsmall = 3))
tbl_footer <- tibble::tribble(
  ~Variable,             ~Type, ~OLS_Beta,            ~OLS_SE, ~`2SLS_Beta`,         ~`2SLS_SE`,
  "N (counties)",        "",      as.character(N_ols),  "",      as.character(N_2sls),   "",
  "Sum of weights",      "",      as.character(sum_w),  "",      as.character(sum_w),  "",
  "R²",                  "",      f_num(r2_ols),      "",      f_num(r2_2sls),       ""
)

#combine
tbl_tbl <- dplyr::bind_rows(
  tbl_bod %>% dplyr::mutate(dplyr::across(dplyr::everything(), as.character)),
  tibble::tibble(Variable = "—", Type = "", OLS_Beta = "", OLS_SE = "", `2SLS_Beta` = "", `2SLS_SE` = ""),
  tbl_footer
)

#render
tbl_ttl <- "Table 1. Examining Slavery's Impact on Contemporary Racial Resentment, Using Soil Suitability for Cotton as an Instrument"

tbl_tbl %>%
  kbl(
    align = c("l","l","r","r","r","r"),
    booktabs = TRUE,
    caption = tbl_ttl,
    col.names = c("Variable","Type","Beta","SE","Beta","SE")
  ) %>%
  add_header_above(c(" " = 2, "OLS" = 2, "2SLS" = 2)) %>%
  kable_classic(full_width = FALSE, html_font = "Times") %>%
  row_spec(nrow(tbl_bod) + 1, extra_css = "border-bottom: 1px solid;") %>% 
  footnote(
    general = paste0(
      "Weighted least squares by county sample size. Two-stage least squares uses soil suitability for cotton as an ",
      "instrument for 1860 slave proportion. Robust (HC3) standard errors reported. ",
      "Alabama omitted as the comparison category for state fixed effects. ",
      "Significance codes: * p < 0.10, ** p < 0.05, *** p < 0.01."
    ),
    threeparttable = TRUE, general_title = ""
  )
```

### 2c. Interpretation: pslave1860

A one unit or 10% increase in the proportion of a county's 1860 population that was enslaved is associated with a 0.666 percentage point increase in anti-Black attitudes today, holding all else constant.

..the relationship isn't statistically significant (95% CI includes 0, p = 0.17). Because of fixed effects for states, this reflects estimates of within-state comparisons.

The coefficient for pslave 1860 is produced from using soil cotton suitability as an instrument.

Notes:

-   This means that cotton suitability was regressed on pslave1860 and all predictors.

-   This assumes that soil cotton suitability affects contemporary levels of county racial resentment through and only through the amount of slavery (measured as proportion of population in that county enslaved in 1860).

-   This assumption could be problematic, if the suitability of soil for growing cotton impacts contemporary racial resentment via other channels.

### 2d. Two-stage approach

```{r echo=FALSE}
#for endogenous variable in the model, regress it on all exogenous independent variables and all instruments. generate predicted values from those first stage regressions.  
mS01 <- lm(pslave1860 ~ cottonsuit
          + ln_coarea00 + rugged + latitude  + longitude + latitude_sq + longitude_sq + water1860 + state.abb_af , data = cleaned_abs, weights = sample.size)

cleaned_abs$pslave_hat <- fitted(mS01)


#Include predicted values as substitute predictors for the original exogenous version.

mS02 <- lm(resent ~ pslave_hat
          + ln_coarea00 + rugged + latitude  + longitude + latitude_sq + longitude_sq + water1860 + state.abb_af, data = cleaned_abs, weights = sample.size)

```

Does two-stage approach get the same as ivreg 2SLS?

```{r, echo=FALSE}
#summary(mS01)
#summary(mS02)
#summary(m02)

c("manual_2sls" = coef(mS02)["pslave_hat"],
  "ivreg_2sls" = coef(m02)["pslave1860"])

```

### 2e. Interpretation: Instrument strength

Looking first at the t value for cottonsuit in the first stage model, the t value is very large (9.707) and p value is less than 0.05 (almost 0, \< 2e-16). The coefficient is 0.408.

Next, I compare the 2sls version of pslave to WLS. The increased coefficient estimate suggests that the OLS was downward-biased + inconsistent, and the effect of pslave1860 attenuated.

I can reject the null at 5% because the Wu-Hausman test is *just* under 0.05. But it's also not that low. However, the F stat from the weak instrument test is far over 10 (94.225) and it's p-value almost 0, suggesting cottonsuit is a strong instrument.

There's no Sargan here because with only one instrument for one endogenous regressor, the model is exactly identified (no over-identification, so no Sargan over-id test is applicable)
