---
title: "mensik_assignment_3"
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: xelatex
---

```{r, setup, echo=FALSE, warning=FALSE, include=FALSE}

library(ggplot2)
library(knitr)
library(kableExtra)
library(rdrobust)
library(mpower)

```

Note: Pretending I am interested in whether experiences with the the CJ system change how people think about policy/politics. I want to know if *being incarcerated* or crossing some threshold of exposure causes people to see the world differently. Pretending that there is such a thing as a 0-100 sentencing score that determines whether someone is sentenced to prison time or parole at 50 (at or greater means prison, less means parole), and that is my effectively arbitrary cutoff turned randomization device. (A bad example in a lot of ways)

So:

-   running/forcing/assignment variable is sentencing score

-   cutoff: where treatment rule switches is 50 points

-   treatment assignment: defendants above 50 get treatment, below do not

-   outcome: policy support

-   key assumption: all traits (both observed and unobserved) other than sentencing score change smoothly with the score, except for treatment

-   estimation: compare predicted outcome (policy support) on either side of 50 (the "jump") using local regressions

```{r, warning=FALSE}

set.seed(15000)

# set up simulation for the running variable
running_var_sent_score <- rnorm(2000, 50, 20) #normally distributed with mean 50, sd 20
running_var_sent_score <- running_var_sent_score[running_var_sent_score >= 0 & running_var_sent_score < 100] #truncated to the [0,100) interval, so observation's treatment status will depend on whether it running var is above or below 50 

# create the treatment assignment -- here is where "defendents whose sentencing scores is less than 50" are assigned to the treatment (incarceration) (D=1), otherwise parole (0). "This is "deterministic discontinuity / the "sharp" RDD setup
D_prison_time <- ifelse(running_var_sent_score > 50, 1, 0)

# simulated outcome (observed y, via the 'true' GDP); added treatment effect of 100 for those above the cutoff + random noise makes it realistic (rnorm)
# baseline policy support is 45, being treated increases support by 100 units ("the jump")
# 0.022 * running_var means outcome increases by 2.2 points for every one unit increase in sentencing score, hold ing all else constant
y_policy_support <- 45 + D_prison_time * 100 + 1.5 * running_var_sent_score - 0.022 * running_var_sent_score^2 + rnorm(length(running_var_sent_score), 0, 100) # plus random noise

plot(running_var_sent_score, y_policy_support, pch = 19, cex = 0.2) 
abline(v = 50, col = "red")






# Bin the data and fit local polynomials... plot (using 50 bins left and right of c_0)
# with p=2 polynomial fitted below and above c_0
# (h = full span of data)
# rdplot bins data to left and right of cutoff sepearately and fits polynomia of order p
# uses 4th degree polynomia and 50 bins on each side (show binned means with fitted lines), visual confirmation of discontinuity
rdplot(y = y_policy_support, x = running_var_sent_score, c = 50, p = 4, nbins = c(50, 50))

# automatic bin N selection
# (selects MSE optimal choice)
rdplot(y = y_policy_support, x = running_var_sent_score, c = 50, binselect = "esmv")

# Illustration: the role of h (varying the bandwith to see the local nature)
rdplot(y = y_policy_support, x = running_var_sent_score, c = 50, binselect = "esmv", h = 50) # h controls how much data around the cutoff is used; small h means more local, less biased, more variance, large h means smoothing, more bias less variance (shows how estimate of the jump depends on h)
rdplot(y = y_policy_support, x = running_var_sent_score, c = 50, binselect = "esmv", h = 30)
rdplot(y = y_policy_support, x = running_var_sent_score, c = 50, binselect = "esmv", h = 10)

# BW selection
summary(rdbwselect(y = y_policy_support, x = running_var_sent_score, c = 50, bwselect = "mserd")) # mserd minimizes mean sq error for the discontinuity
summary(rdbwselect(y = y_policy_support, x = running_var_sent_score, c = 50, bwselect = "msetwo")) # allows different bandwidths left/right of the cutoff



# estimate RDD treatment effect
# p = 1 (locally linear)
views_m1 <- rdrobust(y = y_policy_support, x = running_var_sent_score, c = 50, p = 1, bwselect = "mserd") #rdrobust is main RDD estimator, fits local polynomia regression (here p = 1 or p = 2 ) within the optimal bandwiths


# Simple RDD p = 2
views_m2 <- rdrobust(y = y_policy_support, x = running_var_sent_score, c = 50, p = 2, bwselect = "mserd")
summary(views_m1)
summary(views_m2)
 

# Construct plot (using obs within selected bw)
# xc = x - 50 centers the running variable so that the cutoff is at 0; 
xc <- running_var_sent_score - 50
m <- rdrobust(y = y_policy_support, x = xc, p = 1)
rdplot(y = y_policy_support, x = xc, 
    subset = -m$bws[1,1]<= xc & xc <= m$bws[1,2],
    binselect = "esmv", kernel = "triangular", 
    h = c(m$bws[1,1], m$bws[1,2]), p = 1)

```

The RD effect of 93.706 means that crossing from a sentencing score of less than 50 to 50 increases policy support by 93.706 points, and if I repreated the whole study many times and built 95% confidence intervals each time, 95% of those intervals would contain the true effect of a jump between 56.815 and 124.840 points. The p score is 0 – the effect is significant.

Part 2

Adding a confounder Z that, in first scenario where continuity holds, would need to change gradually with running variable/not be discontinuous at the 50 juncture.

```{r}

Z_noproblem <- running_var_sent_score*0.1 +
  rnorm(length(running_var_sent_score), 0, 500) # plus noise one draw per observation


y_confounded <- 45 + D_prison_time * 100 + 1.5 * running_var_sent_score +
                0.022 * Z_noproblem - 0.015 * running_var_sent_score^2 +
                rnorm(length(running_var_sent_score), 0, 100) #adjusted back


# local linear model

## y_confounded
m_a <- rdrobust(y = y_confounded, x = running_var_sent_score, c = 50, p = 1, bwselect = "mserd")
summary(m_a)


# local quadratic model

## y_confounded
m_c <- rdrobust(y = y_confounded, x = running_var_sent_score, c = 50, p = 2, bwselect = "mserd")
summary(m_c)


## y_confounded plot
rdplot(y = y_confounded, x = running_var_sent_score, c = 50, binselect = "esmv")




### then 

#with dummy to jump at 50, by 50
Z_problem <- 0.1*running_var_sent_score + 50 * ifelse(running_var_sent_score > 50, 1, 0) + 
               rnorm(length(running_var_sent_score), 0, 10)

y_bad_confounded <- 45 + D_prison_time * 100 + 1.5 * running_var_sent_score +
                0.022 * Z_problem - 0.015 * running_var_sent_score^2 +
                rnorm(length(running_var_sent_score), 0, 100)



# local linear model

## y_bad_confounded
m_b <- rdrobust(y = y_bad_confounded, x = running_var_sent_score, c = 50, p = 1, bwselect = "mserd")
summary(m_b)

# local quadratic model

## y_bad_confounded
m_d <- rdrobust(y = y_bad_confounded, x = running_var_sent_score, c = 50, p = 2, bwselect = "mserd")
summary(m_d)

# plot

## y_bad_confounded
rdplot(y = y_bad_confounded, x = running_var_sent_score, c = 50, binselect = "esmv")

```

Re-running both scenarios, the smooth confounder stays closer to the true effect but the problematic/discontinuous jumper exerts negative bias on the effect. The discontinuity contaminates the estimate because it reflects the treatment plus the jump.

Third -- the characteristics of the RDD estimator (e.g., using linear vs higher-order local regression) when the data is noisy. What can you learn about false positive results?

Addint noise to the unproblematically confounded estimator increases the p value, unsurprisingly distorts point estimates.

(tried to set up MC but ran out of time)

```{r}
#| eval: false
#| include: false

#monte carlo test 

set.seed (1)

x <- rnorm(2000, 50, 20)

x <- x[x >= 0 & x <100]

D <- ifelse(x > 50, 1, 0)

c0 <- 50
c <- x -c0

#dgp without true treatment effect

y_generator <- function(sigma_y = 100) {
  45 + 0*D + 1.5*x - 0.022*x^2 + rnorm(length(x), 0, sigma_y)
}

run <- function(sigma_y = 100, h_local = 10){
  y <- gen_y(sigma_y)


# -- glitchy OLS

m_glitch <- lm(y_generator~D +x)
p_glitch <- coef(summary(m_glitch))["D", "P(t)"]

# RD OLS

include <- abs(c)<= h_local
m_rd <- lm(y[include] + c[include] + D[include]:c[include])
p_rd <- coef(summary(m_rd))["D", "P(t)"]

# robust rd

m_robust <- rdrobust(y=y_generator, x=x, c= c0, p = 1, bwselect = "mserd")
p_robust <- m_robust$pv[1]

c(p_glitch=p_glitch, p_rd=p_rd, p_robust=p_robust)
}

repeats <- 1000

try <- replicate(repeats, run(sigma_y = 200))

apply(try < 0.05, 1 ,mean)

```
