---
title: "PS12"
output: 
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

## Instructions

Create a new markdown document where you will answer all questions. Submit (1) your code as an `.Rmd` file, and (2) a rendered html file including only what you wish to present, to the `Problem Set 11` assignment under the Assignments tab in Canvas. The problem set is due *Friday, 25 April, by 10:00 EST*. Name your files `lastname_ps12` (`.html` and `.Rmd`).

All submitted work must be your own.

## Packages

You should install and load the following packages. For `synthdid`, install (if needed) via `devtools`:

```{r echo=TRUE, eval=FALSE}
devtools::install_github("synth-inference/synthdid")
```

```{r echo=TRUE, warning=FALSE}
library(synthdid)
library(did)
library(fixest)
library(plm)
```

## Problem 1

For the first problem, you will use the `california_prop99` dataset from the `synthdid` package. Load the data using `data(california_prop99)`. Use `?california_prop99` to see the data description and the variables available.

a.  Using either the `fixest` or `plm` package, estimate a two-way fixed effects model to test the hypothesis that California's cigarette tax hike in 1989 reduced per-capita cigarette consumption relative to the other control states. HINT: you need to explicitly call `Year` as a factor variable.

b.  Interpret the coefficient on `treated`.

c.  Calculate clustered standard errors by `State` (HINT: the packages have functions for this available) and construct a 95% confidence interval for the effect of the CA tax hike on cigarette consumption.

d.  Now estimate the same model using `lm`, and show you get the same estimate for `treated`.

## Problem 2

For this analysis, you will use the dataset `mpdta` from the package `did`. Load the data using `data(mpdta)`. Use `?mpdta` to see the data description and the variables available.

a.  You first need to recode the `treat` variable so that it is 0 in years before treatment. Use this code below:

```{r echo=TRUE}
mpdta$treat_01 <- ifelse(mpdta$first.treat != 0 & mpdta$year >= mpdta$first.treat, 1, 0)
```

b.  Use `feols` in the `fixest` package to estimate a two-way fixed effects model to test the hypothesis that raising the minimum wage reduces teen employment. Use clustered standard errors by county.

c.  Estimate a dynamic version of this model using the code below, then plot the model object using `iplot`. Is there evidence against the parallel trends assumption?

```{r echo=TRUE}
# estimate dynamic version of the model
m_dynamic  <- feols(lemp ~ sunab(first.treat, year) | countyreal + year,
                    data = mpdta, vcov = cluster ~ countyreal)

iplot(m_dynamic, xlab = "Years since minimum wage increase", ref.line = 0)
```

d.  All evidence considered, would you say there is support for the claim that minimum wage hikes reduced youth employment?
