---
title: "Mensik Experiments PS 4"
author: "Kristina Mensik"
date: "2026-10-04"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(estimatr)
library(tidyverse)
library(randomizr)
library(kableExtra)
library(ggplot2)
library(blockTools)


getwd()

```

Questions 1, 2, 3, 4, 5, 6, 7, 9, 10

# Question 1

## a

Covariates are "supplementary variables that predict outcomes" (G & G
95). They are not required for unbiased estimation. They're used to
improve precision by reducing variance in potential outcomes or
differences between treatment and control observations in OLS, and to
assign blocks in block randomized assignment. In theory, they are
measured before subjects are randomized into treatment and control.

## b

"Disturbance term" refers to the part of a regression that is
unobserved, which is unexplained variation in untreated and treated
outcomes.

## c

Equation 4.2 demonstrates that rescaling the outcome by subtracting a
pretest leads to unbiased estimates of the ATE. Subtracting the rescaled
pretest $cX_j$ instead of the pre-test $X_i$ would....(show it produces
unbaised estiamtes of the ATE).

Equation 4.2 says

$$
E(\hat{ATE}) = E[Y_i - X_i|D_i = 1] - E[Y_i-X_i|D_i = 0]
$$ The rescaled pretest $cX_i$ where $c$ is constant and positive can be
substituted for $X_i$. $$
E(\hat{ATE}) = E[Y_i - cX_i|D_i = 1] - E[Y_i-cX_i|D_i = 0]
$$ So when we expand as in 4.2: $$
E(\hat{ATE}) = E[Y_i|D_i = 1] - E[cX_i|D_i=1] - E[Y_i|D_i = 0] + cE[X_i|Di=0]
$$ ...and rearrange: $$
E(\hat{ATE}) = E[Y_i|D_i = 1] - E[Y_i|D_i = 0] - c(E[X_i|D_i=1]  + E[X_i|Di=0])
$$ ...and because we assume that the expected values of $Y_i$ when
$D_i = 0$ are the same as when $D_i = 1$, we arrive at the same place as
in figure 4.2: $$
 E(\hat{ATE}) = E(Y_i(1))-E[Y_i(0)]
$$

## d

The slope parameter $b$ in equation 4.7 is the change in average
potential outcomes $\mu_{Y(1)} - \mu_{Y(0)}$, also called the ATE.

$$
E(ATE) = E[Y_i(1)] - E[Y_i(0)]
$$ \## e Balanced designs are ideal because they allow for interval
estimation: CI's will be conservative rather than too wide or narrow
from outcome variances with imbalance in the treatment and controls that
produce. They're particularly helpful when there is similar variance in
the outcomes of treatment and control groups.

# Question 2

Puzzles after treatment, recorded number of correctly solved puzzles.
control students given 30 extra minutes a few days later to improve
puzzle-solving skills, treatment group were given rules of thumb; all
then took post-test.

## a.

RI to test the null that pre-test scores are unaffected by treatment
assignment.

```{r}
data_2 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Gendelman_2004.csv")

set.seed(312608170)

names(data_2)

subject <- data_2$subject
treatment <- data_2$treatment
pretest <- data_2$pretest
posttest <- data_2$posttest
improvement <- data_2$improvement

model_2 <- lm_robust(treatment ~ pretest)



fstat_2 <- model_2$fstatistic # f stat of treatment

#fstat_2

ra_2 <- declare_ra(nrow(data_2), m = 9) # random assignment procedure

perm_matrx_2 <- obtain_permutation_matrix(ra_2) #10k randomization permmatrix

estimates_2 <- c() #bucket to store results

for (i in 1:ncol(perm_matrx_2)) {
  data_2$ri <- perm_matrx_2[, i]
  random_inf_2 <- lm_robust(data_2$ri ~ data_2$pretest)
  estimates_2 <- c(estimates_2, random_inf_2$fstatistic[1])
}
```

The reported p-value is:

```{r}
print(mean(estimates_2 >= random_inf_2$fstatistic))
```

We can't reject the null hypothesis that treatment had no effect on
pretest scores because 0.1972 is much higher than 0.05. In this case,
rejecting the null means randomization was successful.

## b

```{r}

dim_2 <- difference_in_means(posttest ~ treatment, data = data_2)
dim_ate_2 <- dim_2$coefficients

```

Treatment's effect on posttest scores is:

```{r}

print(dim_ate_2)

```

```{r}
set.seed(312608170)

null <- 0

dim_estimates_2 <- c()

data_2 <- data_2 %>%
  mutate(
    y0_null = case_when(
      treatment == 0 ~ posttest,
      treatment == 1 ~ posttest - null
    ),
    y1_null = case_when(
      treatment == 1 ~ posttest,
      treatment == 0 ~ posttest + null
    )
  )

unique(data_2$y1_null - data_2$y0_null == null)

```

```{r}

for(i in 1:ncol(perm_matrx_2)){
 
   data_2$ri <- perm_matrx_2[, i] #hypothetical assignment
  
  
  data_2$y_temp <- ifelse(
    data_2$ri == 1,
    data_2$y1_null,
    data_2$y0_null
  )
  
  mean_treated <- mean(data_2$y_temp[data_2$ri == 1]) #DIM
  mean_control <- mean(data_2$y_temp[data_2$ri == 0])
  
  dim_estimates_2[i] <- mean_treated - mean_control #store 
  
}

mean(dim_estimates_2) # should be centered around 0

# CI? 

ci_2 <- quantile(dim_estimates_2, probs = c(0.025, 0.975))

ci_2

```

as a table:

```{r}

knitr::kable(ci_2, caption = "CIs for DIM", col.names = "Est.") 
  
```

## c

D-i-D to estimate effect of treatment on post-test score:

```{r}

did_2 <- difference_in_means(improvement ~ treatment, data = data_2)

did_ate_2 <- did_2$coefficients

print(did_ate_2)

```

95% CI, and compare to interval from b.

```{r}
#CI's from difference_in_means() function

did_ci_2 <- c(
  lower = dim_2$conf.low,
  upper = dim_2$conf.high
)

did_ci_2

knitr::kable(did_ci_2, caption = "CIs for DIM", col.names = "Est.") 

```

The interval is wider than for the DIM estimation of the effect of
treatment on the postscore. This suggests that adjusting for the
pre-test covariate with the D-i-D estimation did not lead to gains in
precision, which is surprising.

# Question 3

This reporting policy does not offer up unbiased estimates because it is
premised on the outcomes -- seems like rather explicit
researcher-induced bias.

# Question 4

20 schools assigned to treatment and 20 schools assigned to controls

```{r}

data_4 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/GerberGreenBook_Chapter4_Table_4_1.csv")

```

## a

```{r}

data_4$obs_y <- ifelse(data_4$D == 1, data_4$y1, data_4$y0) 

#regress treatment on observed outcomes

model_4 <- lm_robust(obs_y ~ D, data = data_4)

summary(model_4)

```

```{r}

dim_4 <- difference_in_means(obs_y ~ D, data = data_4)
dim_ate_4 <- dim_4$coefficients

print(dim_ate_4)
```

```{r}
print(coef(model_4)["D"])
```

Is estimated slope the same as estimated ATE based on DIM

```{r}
mean(data_4$obs_y [data_4$D == 0]) #control group mean should = intercept value
```

Yes.

## b

Treated and untreated outcomes on $X_i$.

```{r}

model_4b <- lm_robust(y0 ~ x, data = data_4)
summary(model_4b)

print(coef(model_4b)["x"])
```

```{r}

model_4b_2 <- lm_robust(y1 ~ x, data = data_4)

summary(model_4b_2)


```

```{r}
print(coef(model_4b_2)["x"])
```

Sum of the slope coefficients:

```{r}
0.9417928 + 0.9420107 
```

. . . is greater than 1, which means the dependent variable needs to be
rescaled so that the outcome is a change score.

## c

```{r}

model_4c <- lm_robust(obs_y ~ D + x, data = data_4)

summary(model_4c)

```

```{r}
print(coef(model_4c)["D"])
```

The results here provide a smaller estimate for D (close to halved). The
DIM approach then produces a smaller treatment effect. The SE is
smaller, so we should be more confident in these results.

## d

```{r}

set.seed(6081709)

declare_4c <- declare_ra(nrow(data_4), m = sum(data_4$D))
declare_4c

```

```{r}
#10k matrix

matrix_4 <- replicate(1000, conduct_ra(declare_4c))

estimates_4 <- rep(NA, 1000)

# dim for every permutation

for (i in 1:ncol(matrix_4)) {
  data_4$ri <- matrix_4[, i]
  
  estimates_4 [i] <- lm_robust(obs_y ~ ri, data = data_4)$coef["ri"]
}

print(round(mean(estimates_4), 2))

```

The two sided p value is:

```{r}
print(mean(abs(estimates_4) >= abs(10.7 )))
```

...less than 0.05 so we are good to reject the sharp null that there is
no effect for any school.

## e

Estimates from part (c):

```{r}
coef_4c<-(coef(model_4c)["D"])

print(coef_4c)
```

RI to eval sharp null, then obtain sampling distribution with 100k
random assignments and for each estimate ATE by regressing Yi on di and
Xi, then interpret.

```{r}

set.seed(6081709)

coef_estimates_4c <- rep(NA, 100000)

for(i in 1:ncol(matrix_4)) {
  
  data_4$ri <- matrix_4[, i]
  
  coef_estimates_4c <- lm(obs_y ~ ri + x, data = data_4)$coef["ri"]
}


print(round(mean(coef_estimates_4c), 2))


```

is not centered around 0!

## f

95% CI estimated with RI:

```{r}

data_4$y0_ci <- ifelse(data_4$D == 0, data_4$obs_y, data_4$obs_y - dim_ate_4)
data_4$y1_ci <- ifelse(data_4$D == 0, data_4$obs_y + dim_ate_4, data_4$obs_y)


set.seed(6081709)

# RI

ate_estimates_4 <- rep(NA, 10000) #init

#ate

for( i in 1:ncol(matrix_4)) {
  
  data_4$ri <- matrix_4[, i]
  
  data_4$y_temp <- data_4$y1_ci*data_4$ri + data_4$y0_ci*(1-data_4$ri)
  
  ate_estimates_4 [i] <- lm_robust(y_temp ~ ri, data = data_4)$coef["ri"]
}


# lower bound

sort_ATEs <- sort(ate_estimates_4)

ci_lower <- sort_ATEs()

```

# Question 5

# Question 6

# Question 7

# Question 9

```{r}
data_9 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Gerber_Green_AAAPSS_2005.csv")
```

## a

The effect of getting phone banked on turnout was positive in Iowa
districts that were not competitive. The effect was positive in the
Michigan districts that were competitive, and negative in those that
were non competitive. Neither was particularly pronounced.

## b 

The composition of treated subjects within each block and differences in
block sizes are obscured in the total ATE.

## c

The ATE estimated by overall Block:

```{r}

block1 <- 0.049487*0.00964
block2 <- 0.1520981*(-0.007829)
block3 <- 0.626616*(-0.01362)
block4<-0.171799*0.008271

ate_9 <- block1+block2+block3+block4

print(ate_9)
```

## d

The OLS weights consider within-block balance and the size of blocks for
control and treatment blocks. Balanced blocks have lower sampling
variability than do imbalanced blocks, so balanced blocks are more
heavily weighted. The OLS weights in 4.3 are more balanced than in part
c.

## e

```{r}

data_9a <- data_9[data_9$strata == 1,]
data_9b <- data_9[data_9$strata == 2,]
data_9c <- data_9[data_9$strata == 3,]
data_9d <- data_9[data_9$strata == 4,]

#strata

data_9a <- ifelse(data_9a$treat2 == 1,
                  sum((data_9a$treat2 == 1)) / nrow(data_9a),
                  sum((data_9a$treat2 == 0 )) / nrow(data_9a))


data_9b <- ifelse(data_9b$treat2 == 1,
                  sum((data_9b$treat2 == 1)) / nrow(data_9b),
                  sum((data_9b$treat2 == 0 )) / nrow(data_9b))

data_9c <- ifelse(data_9c$treat2 == 1,
               sum((data_9c$treat2 == 1)) / nrow(data_9c),
               sum((data_9c$treat2 == 0 )) / nrow(data_9c))

data_9d <- ifelse(data_9d$treat2 == 1,
                  sum((data_9d$treat2 == 1)) / nrow(data_9d),
                  sum((data_9d$treat2 == 0 )) / nrow(data_9d))

data_9_weighted <- rbind(data_9a, data_9b, data_9c, data_9d)
  
#regress after inverting weights

data_9_weighted$weight <- 1/data_9_weighted$q # not working

# would then regress weighted vote on treatment

model_9 <- lm_robust(vote02 ~ treat + strata, weights = weight, data = data_9_weighted)

```

# Question 10

Cluster Random Design for KC voter mobilization.

```{r}

data_10 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Arceneaux_AAAPSSsubset_2005.csv")

```
