---
title: "Mensik Experiments PS 4"
author: "Kristina Mensik"
date: "2026-10-04"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(estimatr)
library(tidyverse)
library(randomizr)
library(kableExtra)
library(ggplot2)
library(blockTools)


getwd()

```

Questions 1, 2, 3, 4, 5, 6, 7, 9, 10

# Question 1

### a

Covariates are "supplementary variables that predict outcomes" (G & G
95). They are not required for unbiased estimation. They're used to
improve precision by reducing variance in potential outcomes or
differences between treatment and control observations in OLS, and to
assign blocks in block randomized assignment. In theory, they are
measured before subjects are randomized into treatment and control.

### b

"Disturbance term" refers to the part of a regression that is
unobserved, which is unexplained variation in untreated and treated
outcomes.

### c

Equation 4.2 demonstrates that rescaling the outcome by subtracting a
pretest leads to unbiased estimates of the ATE. Subtracting the rescaled
pretest $cX_j$ instead of the pre-test $X_i$ would....(show it produces
unbaised estiamtes of the ATE).

Equation 4.2 says

$$
E(\hat{ATE}) = E[Y_i - X_i|D_i = 1] - E[Y_i-X_i|D_i = 0]
$$ The rescaled pretest $cX_i$ where $c$ is constant and positive can be
substituted for $X_i$. $$
E(\hat{ATE}) = E[Y_i - cX_i|D_i = 1] - E[Y_i-cX_i|D_i = 0]
$$ So when we expand as in 4.2: $$
E(\hat{ATE}) = E[Y_i|D_i = 1] - E[cX_i|D_i=1] - E[Y_i|D_i = 0] + cE[X_i|Di=0]
$$ ...and rearrange: $$
E(\hat{ATE}) = E[Y_i|D_i = 1] - E[Y_i|D_i = 0] - c(E[X_i|D_i=1]  + E[X_i|Di=0])
$$ ...and because we assume that the expected values of $Y_i$ when
$D_i = 0$ are the same as when $D_i = 1$, we arrive at the same place as
in figure 4.2: 

$$
 E(\hat{ATE}) = E(Y_i(1))-E[Y_i(0)]
$$



### d

The slope parameter $b$ in equation 4.7 is the change in average
potential outcomes $\mu_{Y1} - \mu_{Y0}$, also called the ATE.

$$
E(ATE) = E[Y_i(1)] - E[Y_i(0)]
$$ 

### e 
Balanced designs are ideal because they allow for interval
estimation: CI's will be conservative rather than too wide or narrow
from outcome variances with imbalance in the treatment and controls that
produce. They're particularly helpful when there is similar variance in
the outcomes of treatment and control groups.

## Question 2

Puzzles after treatment, recorded number of correctly solved puzzles.
control students given 30 extra minutes a few days later to improve
puzzle-solving skills, treatment group were given rules of thumb; all
then took post-test.

### a.

RI to test the null that pre-test scores are unaffected by treatment
assignment.

```{r}
data_2 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Gendelman_2004.csv")

set.seed(312608170)

names(data_2)

subject <- data_2$subject
treatment <- data_2$treatment
pretest <- data_2$pretest
posttest <- data_2$posttest
improvement <- data_2$improvement

model_2 <- lm_robust(treatment ~ pretest)



fstat_2 <- model_2$fstatistic # f stat of treatment

#fstat_2

ra_2 <- declare_ra(nrow(data_2), m = 9) # random assignment procedure

perm_matrx_2 <- obtain_permutation_matrix(ra_2) #10k randomization permmatrix

estimates_2 <- c() #bucket to store results

for (i in 1:ncol(perm_matrx_2)) {
  data_2$ri <- perm_matrx_2[, i]
  random_inf_2 <- lm_robust(data_2$ri ~ data_2$pretest)
  estimates_2 <- c(estimates_2, random_inf_2$fstatistic[1])
}
```

The reported p-value is:

```{r}
print(mean(estimates_2 >= random_inf_2$fstatistic))
```

We can't reject the null hypothesis that treatment had no effect on
pretest scores because 0.1972 is much higher than 0.05. In this case,
rejecting the null means randomization was successful.

### b

```{r}

dim_2 <- difference_in_means(posttest ~ treatment, data = data_2)
dim_ate_2 <- dim_2$coefficients

```

Treatment's effect on posttest scores is:

```{r}

print(dim_ate_2)

```

```{r}
set.seed(312608170)

null <- 0

dim_estimates_2 <- c()

data_2 <- data_2 %>%
  mutate(
    y0_null = case_when(
      treatment == 0 ~ posttest,
      treatment == 1 ~ posttest - null
    ),
    y1_null = case_when(
      treatment == 1 ~ posttest,
      treatment == 0 ~ posttest + null
    )
  )

unique(data_2$y1_null - data_2$y0_null == null)

```

```{r}

for(i in 1:ncol(perm_matrx_2)){
 
   data_2$ri <- perm_matrx_2[, i] #hypothetical assignment
  
  
  data_2$y_temp <- ifelse(
    data_2$ri == 1,
    data_2$y1_null,
    data_2$y0_null
  )
  
  mean_treated <- mean(data_2$y_temp[data_2$ri == 1]) #DIM
  mean_control <- mean(data_2$y_temp[data_2$ri == 0])
  
  dim_estimates_2[i] <- mean_treated - mean_control #store 
  
}

mean(dim_estimates_2) # should be centered around 0

# CI? 

ci_2 <- quantile(dim_estimates_2, probs = c(0.025, 0.975))

ci_2

```

as a table:

```{r}

knitr::kable(ci_2, caption = "CIs for DIM", col.names = "Est.") 
  
```

### c

D-i-D to estimate effect of treatment on post-test score:

```{r}

did_2 <- difference_in_means(improvement ~ treatment, data = data_2)

did_ate_2 <- did_2$coefficients

print(did_ate_2)

```

95% CI, and compare to interval from b.

```{r}
#CI's from difference_in_means() function

did_ci_2 <- c(
  lower = dim_2$conf.low,
  upper = dim_2$conf.high
)

did_ci_2

knitr::kable(did_ci_2, caption = "CIs for DIM", col.names = "Est.") 

```

The interval is wider than for the DIM estimation of the effect of
treatment on the postscore. This suggests that adjusting for the
pre-test covariate with the D-i-D estimation did not lead to gains in
precision, which is surprising.

## Question 3

This reporting policy does not offer up unbiased estimates because it is
premised on the outcomes -- seems like rather explicit
researcher-induced bias.

## Question 4

20 schools assigned to treatment and 20 schools assigned to controls

```{r}

data_4 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/GerberGreenBook_Chapter4_Table_4_1.csv")

```

### a

```{r}

data_4$obs_y <- ifelse(data_4$D == 1, data_4$y1, data_4$y0) 

#regress treatment on observed outcomes

model_4 <- lm_robust(obs_y ~ D, data = data_4)

summary(model_4)

```

```{r}

dim_4 <- difference_in_means(obs_y ~ D, data = data_4)
dim_ate_4 <- dim_4$coefficients

print(dim_ate_4)
```

```{r}
print(coef(model_4)["D"])
```

Is estimated slope the same as estimated ATE based on DIM

```{r}
mean(data_4$obs_y [data_4$D == 0]) #control group mean should = intercept value
```

Yes.

### b

Treated and untreated outcomes on $X_i$.

```{r}

model_4b <- lm_robust(y0 ~ x, data = data_4)
summary(model_4b)

print(coef(model_4b)["x"])
```

```{r}

model_4b_2 <- lm_robust(y1 ~ x, data = data_4)

summary(model_4b_2)


```

```{r}
print(coef(model_4b_2)["x"])
```

Sum of the slope coefficients:

```{r}
0.9417928 + 0.9420107 
```

. . . is greater than 1, which means the dependent variable needs to be
rescaled so that the outcome is a change score.

### c

```{r}

model_4c <- lm_robust(obs_y ~ D + x, data = data_4)

summary(model_4c)

```

```{r}
print(coef(model_4c)["D"])
```

The results here provide a smaller estimate for D (close to halved). The
DIM approach then produces a smaller treatment effect. The SE is
smaller, so we should be more confident in these results.

### d

```{r}

set.seed(6081709)

declare_4c <- declare_ra(nrow(data_4), m = sum(data_4$D))
declare_4c

```

```{r}
#10k matrix

matrix_4 <- replicate(1000, conduct_ra(declare_4c))

estimates_4 <- rep(NA, 1000)

# dim for every permutation

for (i in 1:ncol(matrix_4)) {
  data_4$ri <- matrix_4[, i]
  
  estimates_4 [i] <- lm_robust(obs_y ~ ri, data = data_4)$coef["ri"]
}

print(round(mean(estimates_4), 2))

```

The two sided p value is:

```{r}
print(mean(abs(estimates_4) >= abs(10.7 )))
```

...less than 0.05 so we are good to reject the sharp null that there is
no effect for any school.

### e

Estimates from part (c):

```{r}
coef_4c<-(coef(model_4c)["D"])

print(coef_4c)
```

RI to eval sharp null, then obtain sampling distribution with 100k
random assignments and for each estimate ATE by regressing Yi on di and
Xi, then interpret.

```{r}

set.seed(6081709)

coef_estimates_4c <- rep(NA, 100000)

for(i in 1:ncol(matrix_4)) {
  
  data_4$ri <- matrix_4[, i]
  
  coef_estimates_4c <- lm(obs_y ~ ri + x, data = data_4)$coef["ri"]
}


print(round(mean(coef_estimates_4c), 2))


```

is not centered around 0!

### f

95% CI estimated with RI:

```{r}

data_4$y0_ci <- ifelse(data_4$D == 0, data_4$obs_y, data_4$obs_y - dim_ate_4)
data_4$y1_ci <- ifelse(data_4$D == 0, data_4$obs_y + dim_ate_4, data_4$obs_y)


set.seed(6081709)

# RI

ate_estimates_4 <- rep(NA, 10000) #init

#ate

for( i in 1:ncol(matrix_4)) {
  
  data_4$ri <- matrix_4[, i]
  
  data_4$y_temp <- data_4$y1_ci*data_4$ri + data_4$y0_ci*(1-data_4$ri)
  
  ate_estimates_4 [i] <- lm_robust(y_temp ~ ri, data = data_4)$coef["ri"]
}


# lower bound

sort_ATEs <- sort(ate_estimates_4)

ci_lower <- sort_ATEs()

```

### Question 5

```{r, echo = T, include = T, warning = F}

data_5 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/GerberGreenBook_Chapter4_Table_4_1.csv")


# restriction that a regression of $d_i$ on $X_i$ (the pre-test) generates and F stat whose reported p value is greater that 0.05. Ie were assigned $d_i$  significantly predicted by $X-i$, RI would be conducted again until $d_i$ met criterion.

#fstat_5 <- lm_robust()

restricted_assignment_5 <- function (d, x, m) {
  while (TRUE) {
   
    Z <- complete_ra(nrow(d), m)
    f <- lm_robust(as.formula(paste("Z ~ ", x)), d)$fstatistic 

  if (1 - pf(q = f[1], df1 = f[2], df2 = f[3]) > 0.05) {
  return(Z)
    }
  }
}

set.seed(6081709)

#  randomization (restricted) iterations
sim_matrix_5 <- replicate(10000, 
                           restricted_assignment_5(d = data_5, x = "x", 
                                         m = sum(data_5$D)))

# f stat where reported p val greater than 0.05?

fp <- c()

for (i in 1:ncol(sim_matrix_5)) {
  data_5$d_temp = sim_matrix_5[, i]
  f <- lm_robust(d_temp ~ x, data_5)$fstatistic
  fp[i] <- 1 - pf(q = f[1], df1 = f[2], df2 = f[3])
}

min(fp)

```

The min value of the created vector a p vals is just above the threshhold in a generous read (0.05005606 -> 0.05006 > 0.05000). I'm not sure if the hypothetical researcher would but I'll take this as sufficient to proceed.


#### a

```{r, echo = T, include = T, warning = F}

# about inverse probability weights

# already made matrix of treatment assignments to calculate the probabilities (matrix_5)

# can I use obtain conditional probabilities? 

declare_5 <- declare_ra(permutation_matrix = sim_matrix_5)
declare_5 #declared conditional probabilities procedure

data_5$weights <- 1/(
  obtain_condition_probabilities(declare_5, data_5$D)
)

# weights in the control and treatment groups should vary 
print(sd(data_5$weights[data_5$D == 0]))
print(sd(data_5$weights[data_5$D == 1]))

```

Weights vary by 0.024 SDs in the control group and by 0.018 in the treatment group. 

#### b

```{r, echo = T, include = T, warning = F}

#  actual DIM, weighted

data_5$obs_y <- ifelse(data_5$D == 1, data_5$y1, data_5$y0)

dim_5 <- difference_in_means(obs_y ~ D, weights = weights, data = data_5)

ate_dim_5 <- dim_5$coefficients

print(ate_dim_5)

## RI 

set.seed(6081709)

dim_estimates_5 <- rep(NA, 10000)

# for every permutation, get the DIM

for (i in 1:ncol(sim_matrix_5)) {
  data_5$ri <- sim_matrix_5[, i]

    dim_estimates_5 [i] <- lm_robust(obs_y ~ ri, 
                                      weights = weights, 
                                      data = data_5)$coef["ri"]

}

print(round(mean(dim_estimates_5))) # is center on 0

print(mean(abs(dim_estimates_5) >= abs(ate_dim_5))) # p-value 

```

The p-value of 0.0045 lets us reject the sharp null. 

#### (c)

```{r, echo = T, include = T, warning = F}

# actual coefficients (weighted, covariate-adjusted)

model_5c <- lm_robust(obs_y ~ D + x, 
                                      weights = weights,
                                      data = data_5)

summary(model_5c)

coef_5c <- model_5c$coef["D"]

print(coef_5c)

# do RI

set.seed(6081709)

coef_estimates_5c <- rep(NA, 10000)

# estimating covariate-adjusted regression 

for (i in 1:ncol(sim_matrix_5)) {
  data_5$ri <- sim_matrix_5[, i]

  coef_estimates_5c [i] <- lm_robust(obs_y ~ ri + x, 
                                      weights = weights,
                                      data = data_5)$coef["ri"]

}

# checking that the distribution is centred around 0

print(round(mean(coef_estimates_5c)))

# calculating the p-value 

print(mean(abs(coef_estimates_5c) >= abs(coef_5c))) 

```

Here we get a two-sided p-val of 0.0028 which is less than 0.05 we can reject the sharp null (no effect for any school)


#### d

Sampling distribution comparison: 5b and c to 4 d and e (latter assumed unrestricted randomization).

```{r, echo = T, include = T, warning = F}

# b and c were difference in means
# data_5_bc

data_5b5c <- data.frame(
  value = c(coef_estimates_5c, dim_estimates_5),
  group = factor(c(rep("Unrestricted", length(coef_estimates_5c)), 
                   rep("Restricted", length(dim_estimates_5))))
)

library(ggplot2)

ggplot(data_5b5c, aes(x = value, color = group)) +
  geom_density (alpha = 0.5, linewidth = 1) +
  scale_fill_manual(values = c("darkgreen", "lightblue")) +
  labs(title = "Sampling Distributions: DIM",
       x = "Value", y = "Count", color = "Group") +
  theme_minimal()


# 4 d and e (covariate adjusted regression)

data_4d4e <- data.frame(
  value = c(q4e_coef_estimates, q5c_coef_estimates),
  group = factor(c(rep("Unrestricted", length(q4e_coef_estimates)), 
                   rep("Restricted", length(q5c_coef_estimates))))
)

library(ggplot2)

ggplot(q5d_df2, aes(x = value, color = group)) +
  geom_density (alpha = 0.5, linewidth = 1) +
  scale_fill_manual(values = c("darkgreen", "lightblue")) +
  labs(title = "Samplign Distributions: Unrestricted Randomization",
       x = "Value", y = "Count", color = "Group") +
  theme_minimal()

```

The sampling distributions under the null hypothesis are slightly tighter around 0 when randomisation was restricted, compared to when it was unrestricted. However, this difference is marginal and not particularly visually apparent — especially in the cases of 4e and 5c — likely because unrestricted randomisation was run 100,000 times, whereas the restricted scenarios used only 10,000 simulations.

# Question 6

#### a

```{r, echo = T, include = T, warning = F}

data_6 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/OBrien_Patsiorkovski_ICPSR_1999.csv")

names(data_6)
# adding ids

data_6$id <- seq_len(nrow(data_6))

data_6$group6 <- as.factor(data_6$group6) #make social class variable factor variable

# predict post-intervention by regressing variables

model_6a <- lm(index96 ~ sexresp6 + memberc6 + group6 + index95, 
                data_6)

summary(model_6a)

```

The 1995 evaluations of national conditions seem to most strongly predict the following year's evaluations. The $R^2$ is 0.39 (rounded to two decimal places), indicating that about 39% of the variation in the 1996 evaluations is explained by our model.

#### (b)

```{r, echo = T, include = T, warning = F}


set.seed(6081709)

# convert each level of the factor variable into binary variable

matrix_6 <- model.matrix(~ sexresp6 + memberc6 + group6 + index96 + index97 
                          + id, 
                          data = data_6)

# removing the intercept column

matrix_6 <- matrix_6 [, -1]

head(matrix_6)

# blocking! 

blocks_6 <- block(matrix_6, n.tr = 42, id.vars = "id", block.vars = setdiff(names(matrix_6), "id"))

blocks_6

data_6_blocked <- as.data.frame(matrix_6)

data_6_blocked$block <- createBlockIDs(blocks_6, data_6_blocked, id.var = "id")

head(data_6_blocked)

# checking number of units in each block 

block_counts <- as.data.frame(table(data_6_blocked$block))
names(block_counts) <- c("block", "n")
block_counts

# conducting block random assignment 

declaration_6 <- declare_ra(nrow(data_6_blocked), m = 21, blocks = data_6_blocked$block)
declaration_6

# to check for covariate balance 

# obtaining permutation matrix

perm_matrix_6 <- obtain_permutation_matrix(declaration_6)

# conducting repeated f-tests 

repeated_f_tests <- c()

for (i in 1:ncol(perm_matrix_6)) {
  data_6_blocked$d_temp = perm_matrix_6[, i]
  f <- lm_robust(d_temp ~ sexresp6 + memberc6 + group62 + group63 + group64 + index96, 
                 data = data_6_blocked)$fstatistic
  repeated_f_tests[i] <- 1 - pf(q = f[1], df1 = f[2], df2 = f[3])
}

hist(repeated_f_tests)

```

This distribution of p-values indicates that the groups are not systematically imbalanced across the control and treatment groups with regard to our covariates of interest.

#### (c)

```{r, echo = T, include = T, warning = F}

# setting seed

set.seed(6081709)

# running a for loop to generate sampling distributions - block ra

# creating empty vector to store results of for loop

ate_estimates_6c <- rep(NA, 10000)

# estimating ATE 

for (i in 1:ncol(perm_matrix_6)) {
  data_6_blocked$ri <- perm_matrix_6[, i]

  ate_estimates_6c [i] <- lm(index97 ~ ri, data = data_6_blocked)$coef["ri"]

}

# checking that the distribution is centred around 0

print(round(mean(ate_estimates_6c), 2))


# running a for loop to generate sampling distributions - complete ra

# conducting complete random assignment 

declaration_6c <- declare_ra(nrow(data_6), m = nrow(data_6)/2)
declaration_6c

# obtaining permutation matrix

perm_matrix_6c <- obtain_permutation_matrix(declaration_6c)

# creating empty vector to store results of for loop

ate_estimates_6c2 <- rep(NA, 10000)

# estimating ATE

for (i in 1:ncol(perm_matrix_6c)) {
  data_6$ri <- perm_matrix_6c[, i]

  ate_estimates_6c2 [i] <- lm(index97 ~ ri, data = data_6)$coef["ri"]

}

# checking that the distribution is centred around 0

print(round(mean(ate_estimates_6c2), 2))

# comparing sampling distributions under block and complete ra

data_6c <- data.frame(
  value = c(ate_estimates_6c, ate_estimates_6c2),
  group = factor(c(rep("Block RA", length(ate_estimates_6c)), 
                   rep("Complete RA", length(ate_estimates_6c2))))
)

library(ggplot2)

ggplot(data_6c, aes(x = value, color = group)) +
  geom_density (alpha = 0.5, linewidth = 1) +
  scale_fill_manual(values = c("forestgreen", "lightblue")) +
  labs(title = "Sampling Distributions (Block and Complete Randomisation)",
       x = "Value", y = "Count", color = "Group") +
  theme_minimal()

```

The sampling distribution of the ATE under block random assignment is narrower and more clustered around the 'true' ATE of 0, than the corresponding sampling distribution under complete random assignment.

#### d

```{r, echo = T, include = T, warning = F}

# setting seed

set.seed(6081709)

# running for loop to generate distribution under covariate adjusted regression

# creating empty vector to store results of for loop

ate_estimates_6_d <- rep(NA, 10000)

# estimating ATE

for (i in 1:ncol(perm_matrix_6c)) {
  data_6$ri <- perm_matrix_6c[, i]
  
  ate_estimates_6_d [i] <- lm(index97 ~ ri + 
                              sexresp6 + 
                              memberc6 + 
                              group6 + 
                              index96, data = data_6)$coef["ri"]

}

# checking that the distribution is centred around 0

print(round(mean(ate_estimates_6_d)))

# comparing sampling distributions under block ra and covariate adjustment

data_6c2 <- data.frame(
  value = c(ate_estimates_6c, ate_estimates_6_d),
  group = factor(c(rep("Block RA", length(ate_estimates_6c)), 
                   rep("Covariate Adjustment", length(ate_estimates_6_d))))
)

library(ggplot2)

ggplot(data_6c2, aes(x = value, color = group)) +
  geom_density (alpha = 0.5, linewidth = 1) +
  scale_fill_manual(values = c("forestgreen", "lightblue")) +
  labs(title = "Sampling Distributions 
       (Block Randomisation and Covariate Adjustment)",
       x = "Value", y = "Count", color = "Group") +
  theme_minimal()

```

The sampling distribution of the ATE under covariate adjustment is more clustered around the 'true' ATE of 0, than the corresponding sampling distribution under block random assignment.

## Question 7

### a

```{r, echo = T, include = T, warning = F}

data_7 <- data.frame(
          col1 = c("A", "B", "C", "D"),
          col2 = c(1, 0, 2, 5),
          col3 = c(2, 3, 2, 5))

colnames(data_7) <- c("Subject", "Y0", "Y1")

delcaration_7 <- declare_ra(nrow(data_7), m = 2)
delcaration_7

# creating the matrix of unique perms mechanically 

perms_7 <- combn(4, 2, simplify = FALSE)

n_perms_7 <- length(perms_7)
D <- matrix(0, nrow = n_perms_7, ncol = 4)

for (i in seq_len(n_perms_7)) {
  treated <- perms_7[[i]]
  D[i, treated] <- 1
}

D_7 <- t(D)

# running a for loop to assess sampling variance of DIM estimator
# I use the standard R var function as we have no reason 
# to believe this is the population

set.seed(6081709)

# creating empty vector to store results of for loop

dim_estimates_7 <- c()

# estimating dim 

for (i in 1:ncol(D_7)) {
  data_7$ri <- D_7[, i]
  
  treated_mean <- mean(data_7$`Y1`[data_7$ri == 1])
  control_mean <- mean(data_7$`Y0`[data_7$ri == 0])
  
  dim <- treated_mean - control_mean
  dim_estimates_7 <- c(dim_estimates_7, dim)
}

print(var(dim_estimates_7))

```
Sampling Variance is 
3.4 

#### b

```{r, echo = T, include = T, warning = F}

# I will run randomisations separately for each blocking scheme, and then cbind

# creating each block as its own dataset

block_options <- list(
  list(c("A","B"), c("C","D")),
  list(c("A","C"), c("B","D")),
  list(c("A","D"), c("B","C"))
)

get_block_data <- function(block_options, data) {
  subjects_ordered <- unlist(block_options)
  block_data <- data[match(subjects_ordered, data$Subject), ]
    block_data$Block <- rep(1:length(block_options), each = 2)
  return(block_data)
}

block_datasets <- lapply(block_options, get_block_data, data = data_7)

names(block_datasets) <- paste0("BlockOption", 1:3)

BlockOption1 <- block_datasets$BlockOption1
BlockOption2 <- block_datasets$BlockOption2
BlockOption3 <- block_datasets$BlockOption3

# randomising within blocking options - 1

delcaration_71 <- declare_ra(
  N = nrow(BlockOption1), 
  blocks = BlockOption1$Block,
  block_m = rep(1, length(unique(BlockOption1$Block))) 
)
delcaration_71

# randomising within blocking options - 2

delcaration_72 <- declare_ra(
  N = nrow(BlockOption2), 
  blocks = BlockOption2$Block,
  block_m = rep(1, length(unique(BlockOption2$Block))) 
)
delcaration_72

# randomising within blocking options - 3

delcaration_73 <- declare_ra(
  N = nrow(BlockOption3), 
  blocks = BlockOption3$Block,
  block_m = rep(1, length(unique(BlockOption3$Block))) 
)
delcaration_73

library(randomizr)

# extracting assignment vectors for each blocking optioon - 1 

assignment1 <- obtain_permutation_matrix(delcaration_71)

# extracting assignment vectors for each blocking option - 2 

assignment2 <- obtain_permutation_matrix(delcaration_72)

# extracting assignment vectors for each blocking option - 3

assignment3 <- obtain_permutation_matrix(delcaration_73)

# combining to create permmatrix of unique random assignments

perm_matrix_7b <- cbind(assignment1, assignment2, assignment3)

# running a for loop to assess sampling variance of DIM estimator
# I use the standard R var function as we have no reason 
# to believe this is the population

# setting seed

set.seed(6081709)

# creating empty vector to store results of for loop

dim_estimates_7b <- c()

# estimating dim 

for (i in 1:ncol(perm_matrix_7b)) {
  data_7$ri <- perm_matrix_7b[, i]
  
  treated_mean <- mean(data_7$`Y1`[data_7$ri == 1])
  control_mean <- mean(data_7$`Y0`[data_7$ri == 0])
  
  dim <- treated_mean - control_mean
  dim_estimates_7b <- c(dim_estimates_7b, dim)
}

print(var(dim_estimates_7b))

```

#### (c)



# Question 9

```{r}
data_9 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Gerber_Green_AAAPSS_2005.csv")
```

## a

The effect of getting phone banked on turnout was positive in Iowa
districts that were not competitive. The effect was positive in the
Michigan districts that were competitive, and negative in those that
were non competitive. Neither was particularly pronounced.

## b 

The composition of treated subjects within each block and differences in
block sizes are obscured in the total ATE.

## c

The ATE estimated by overall Block:

```{r}

block1 <- 0.049487*0.00964
block2 <- 0.1520981*(-0.007829)
block3 <- 0.626616*(-0.01362)
block4<-0.171799*0.008271

ate_9 <- block1+block2+block3+block4

print(ate_9)
```

## d

The OLS weights consider within-block balance and the size of blocks for
control and treatment blocks. Balanced blocks have lower sampling
variability than do imbalanced blocks, so balanced blocks are more
heavily weighted. The OLS weights in 4.3 are more balanced than in part
c.

## e

```{r}

data_9a <- data_9[data_9$strata == 1,]
data_9b <- data_9[data_9$strata == 2,]
data_9c <- data_9[data_9$strata == 3,]
data_9d <- data_9[data_9$strata == 4,]

#strata

data_9a <- ifelse(data_9a$treat2 == 1,
                  sum((data_9a$treat2 == 1)) / nrow(data_9a),
                  sum((data_9a$treat2 == 0 )) / nrow(data_9a))


data_9b <- ifelse(data_9b$treat2 == 1,
                  sum((data_9b$treat2 == 1)) / nrow(data_9b),
                  sum((data_9b$treat2 == 0 )) / nrow(data_9b))

data_9c <- ifelse(data_9c$treat2 == 1,
               sum((data_9c$treat2 == 1)) / nrow(data_9c),
               sum((data_9c$treat2 == 0 )) / nrow(data_9c))

data_9d <- ifelse(data_9d$treat2 == 1,
                  sum((data_9d$treat2 == 1)) / nrow(data_9d),
                  sum((data_9d$treat2 == 0 )) / nrow(data_9d))

data_9_weighted <- rbind(data_9a, data_9b, data_9c, data_9d)
  
#regress after inverting weights

data_9_weighted$weight <- 1/data_9_weighted$q # not working

# would then regress weighted vote on treatment

model_9 <- lm_robust(vote02 ~ treat + strata, weights = weight, data = data_9_weighted)

```

# Question 10

Cluster Random Design for KC voter mobilization.

```{r}

data_10 <- read.csv("/Users/kom11/Documents/GitHub/regression-study/Experiments/doi-10.60600-yu-vuh8kd/Arceneaux_AAAPSSsubset_2005.csv")

```
